---
title: "Simplifying ggplot2 code by doing nothing"
excerpt: "The ggplot2 version of multiplying by 1"
tags: ''
---

```{r setup, include = FALSE}
# knitr::opts_chunk$set(echo = TRUE)
```

Recently, I joined the development team for
[bayesplot](http://mc-stan.org/bayesplot/), an R package by the Stan team for
quickly plotting Bayesian models. Because plotting the results of Bayesian
models in ggplot2 is a [recurring]((/visualizing-uncertainty-rstanarm/)
[topic](/plotting-partial-pooling-in-mixed-effects-models/)
[here](/bayesian-fisher-exact-test/), it was a natural fit. I have been cleaning
up and simplifying some of the plotting code in the package, so in this post, I
describe one of the techniques I use in the package: **avoid if-branches by
sometimes plotting nothing**.

## Warm-up example

Let's start with a non-plotting example. We write a function
that takes a list of counts of people and returns the total number of people.
That sum, however, is controlled by some of the function's arguments:

```{r}
head_count <- function(data, staff = FALSE, faculty = FALSE) {
  total <- data[["students"]]
  
  if (staff) {
    total <- total + data[["staff"]]
  }
  
  if (faculty) {
    total <- total + data[["faculty"]]
  }
  
  total
}

data <- list(students = 10, staff = 3, faculty = 2)
head_count(data)
head_count(data, staff = TRUE)
head_count(data, faculty = TRUE)
```

The function as it's written works fine... but the if-branches get in the
way too. The main job of the function is to assemble something (a sum) from
pieces of data, but that assembly is split across the if-branches. The `total`
is updated in-place twice (`total <- total +`). I would rather combine all the
data in one fell swoop.

One way to move the data-assembly out of these if-branches is to make _the data_
conditional. That is, instead of doing the addition in the `if` statements, we
do all the addition at once but just add 0 if a piece of data isn't
needed. This technique yields simplifies the function definition:

```{r}
head_count <- function(data, staff = FALSE, faculty = FALSE) {
  staff <- if (staff) data[["staff"]] else 0
  faculty <- if (faculty) data[["faculty"]] else 0
  
  data[["students"]] + staff + faculty
}

head_count(data)
head_count(data, staff = TRUE)
head_count(data, faculty = TRUE)
```

For operations where we combine things to get a new thing, there's often an
identity element: a value we can plug in as a placeholder value without changing
the results. Some examples:

```{r}
# string concatenation
paste0("a", "")

# multiplication
pi * 1

# combining things
c(1, 2, 3, c())

# data-frame binding
rbind(head(iris), list())

# magrittr pipes
f <- function(x, ...) x
1:2 %>% f()
```

Below, I demonstrate two examples of how I applied a similar strategy in
bayesplot. In ggplot2, we assemble plots by adding layers and options to a plot
by stringing together statements with `+`. To handle conditional plotting
elements and construct the plot all in one pass, I took advantage of ways to
adding nothing to a ggplot2 plot.[^monoids]

## Interval plots

Here are some MCMC samples for a Bayesian model. We want to create a plot that
shows the 90% and 50% intervals for each parameter estimate. We first use a
helper function to clean up the data to prepare it for plotting. (At the time of
writing, `mcmc_intervals_data()` is not yet in the CRAN version of bayesplot.)

```{r}
library(bayesplot)
library(ggplot2)
theme_set(theme_grey())

interval_data <- shinystan::eight_schools@posterior_sample %>% 
  mcmc_intervals_data(pars = "mu", regex_pars = "theta") %>% 
  mutate_if(is.numeric, round, 2) %>% 
  mutate()

interval_data
```

Below is the basic plot we want to create. We use `aes_(y = ~ parameter)`
because in R package programming, writing `aes(y = parameter)` would raise an
undefined global variable warning during CRAN checks. The formula form "quotes"
the variable name, so it doesn't appear as a global variable.

```{r}
ggplot(interval_data) + 
  aes_(y = ~ parameter, yend = ~ parameter) +
  geom_segment(aes_(x = ~ ll, xend = ~ hh), size = 1) + 
  geom_segment(aes_(x = ~ l, xend = ~ h), size = 2) + 
  scale_y_discrete(limits = rev(interval_data$parameter)) + 
  labs(x = NULL, y = NULL)
```

We are going to wrap this code in an R function along with two annotation
options: 

1. whether to draw point estimates (medians) over the intervals 
2. whether to draw a vertical reference line at _x_ = 0

Here is how we might write the function with branching code. 

```{r tests-1, fig.align = "default"}
plot_intervals <- function(data, draw_points = TRUE, draw_ref_line = TRUE,
                           line_position = 0) {
  p <- ggplot(data) + 
    aes_(y = ~ parameter, yend = ~ parameter) 
  
  if (draw_ref_line) {
    p <- p + 
      geom_vline(xintercept = line_position, size = 2, color = "white")
  }
  
  p <- p +
    geom_segment(aes_(x = ~ ll, xend = ~ hh), size = 1) + 
    geom_segment(aes_(x = ~ l, xend = ~ h), size = 2)
  
  if (draw_points) {
    p <- p + geom_point(aes_(x = ~ m), size = 3)
  }
  
   p + 
     scale_y_discrete(limits = rev(data$parameter)) + 
     labs(x = NULL, y = NULL)
}

plot_intervals(interval_data)
plot_intervals(interval_data, draw_points = FALSE)
plot_intervals(interval_data, draw_ref_line = FALSE)
```

As we can see, the plot is built up incrementally throughout the function.
The plot object `p` is updated three times (the `p <- p +` parts). This makes
more difficult to follow the function when reading it and increases the chance
that we might do something wrong when working on this code.

We can improve the readability by using empty or placeholder elements to remove
the plot updates inside of the if-branches.

## Do nothing by plotting an empty dataframe

ggplot2 is built to work on dataframe columns, and if there is no data in a
column, it will quietly plot nothing. Therefore, one way to do nothing is to run
the normal `geom_` plotting function but on an empty dataframe. Here, we apply
that strategy for the point estimates by toggling between the original data or
zero-row alternative in the first line.

```{r test-2, out.width="50%", fig.show='hold', fig.align = "default"}
plot_intervals <- function(data, draw_points = TRUE, draw_ref_line = TRUE,
                           line_position = 0) {
  maybe_points <- if (draw_points) data else data[0, ]
  
  p <- ggplot(data) + 
    aes_(y = ~ parameter, yend = ~ parameter) 
  
  if (draw_ref_line) {
    p <- p + 
      geom_vline(xintercept = line_position, size = 2, color = "white")
  }
  
  p +
    geom_segment(aes_(x = ~ ll, xend = ~ hh), size = 1) + 
    geom_segment(aes_(x = ~ l, xend = ~ h), size = 2) + 
    geom_point(aes_(x = ~ m), data = maybe_points, size = 3) + 
    scale_y_discrete(limits = rev(data$parameter)) + 
    labs(x = NULL, y = NULL)
}

plot_intervals(interval_data)
plot_intervals(interval_data, draw_points = FALSE)
plot_intervals(interval_data, draw_ref_line = FALSE)
```

This is a marked improvement over the original, although

## Do nothing with `geom_blank`

Another way to go nothing is to add `geom_blank()`. In fact, this is what
ggplot2 does when we print a ggplot2 object without any data.

```{r}
ggplot(interval_data) + 
  aes(x = m, y = parameter)
```

When we inspect the last plot that was displayed, we see `geom_blank()` is one of the layers:

```{r}
last_plot()[["layers"]]
```

As a first pass for technique, we toggle between the plotting functions
`geom_vline()` and `geom_blank()`. This version simplifies the plot
construction to a single stream of plot additions.

```{r test-3, out.width="30%", fig.show='hold', fig.align = "default"}
plot_intervals <- function(data, draw_points = TRUE, draw_ref_line = TRUE,
                           line_position = 0) {
  maybe_points <- if (draw_points) data else data[0, ]
  geom_maybe_vline <- if (draw_ref_line) geom_vline else geom_blank
    
  ggplot(data) + 
    aes_(y = ~ parameter, yend = ~ parameter) +
    geom_maybe_vline(xintercept = line_position, size = 2, color = "white") +
    geom_segment(aes_(x = ~ ll, xend = ~ hh), size = 1) + 
    geom_segment(aes_(x = ~ l, xend = ~ h), size = 2) + 
    geom_point(aes_(x = ~ m), data = maybe_points, size = 3) + 
    scale_y_discrete(limits = rev(data$parameter)) + 
    labs(x = NULL, y = NULL)
}
plot_intervals(interval_data)
plot_intervals(interval_data, draw_ref_line = FALSE)
```

But it also issues a warning when it toggles to `geom_blank()`. That's annoying. 

For bayesplot, I wrote a helper called `geom_ignore()` as a version of `geom_blank()` that ignores any input arguments. The function recieves any number of arguments in the `...` argument, but note that that function does nothing with those dots. It just ignores them.

```{r}
geom_ignore <- function(...) {
  geom_blank(mapping = NULL, data = NULL,
             show.legend = FALSE, inherit.aes = FALSE)
}
```

For completeness, here is the final form using `geom_ignore()`.

```{r test-4, out.width="30%", fig.show='hold', fig.align = "default"}
plot_intervals <- function(data, draw_points = TRUE, draw_ref_line = TRUE,
                           line_position = 0) {
  maybe_points <- if (draw_points) data else data[0, ]
  geom_maybe_vline <- if (draw_ref_line) geom_vline else geom_ignore
    
  ggplot(data) + 
    aes_(y = ~ parameter, yend = ~ parameter) +
    geom_maybe_vline(xintercept = line_position, size = 2, color = "white") +
    geom_segment(aes_(x = ~ ll, xend = ~ hh), size = 1) + 
    geom_segment(aes_(x = ~ l, xend = ~ h), size = 2) + 
    geom_point(aes_(x = ~ m), data = maybe_points, size = 3) + 
    scale_y_discrete(limits = rev(data$parameter)) + 
    labs(x = NULL, y = NULL)
}

plot_intervals(interval_data)
plot_intervals(interval_data, draw_ref_line = FALSE)
```


### A final note on `NULL` and why I avoid it

Adding `NULL` to a ggplot doesn't change it (e.g., `last_plot() + NULL`), but I
avoid this technique. First, it seems like an edge-case: We are normally not
supposed to add `NULL` to things, so we're asking for trouble. (In contrast, we are supposed to add `geom_blank()` to things, so no problem there.) Second, it
doesn't seem future-proof. We might imagine a later stricter version of ggplot2
where adding `NULL` raises an error. Then all the code that relies on _this one
weird trick_ will break.



[^monoids]: This technique of using an empty element with combiner functions was
inspired by the concept of [monoids](https://en.wikipedia.org/wiki/Monoid) which
come up in functional programming. Monoids have a stricter formal definition,
and ggplot2 are most certainly not monoids because everything must be added onto
an initial `ggplot()` object.



