---
title: Draft post (2018-10-18)
excerpt: ''
tags: ''
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Start by reading over the [article on extending ggplot2](https://ggplot2.tidyverse.org/articles/extending-ggplot2.html).

```{r}
summarise_by_xn <- function(data, summary, ...) {
  summary <- plyr::ddply(data, c("group", "x"), summary, ...)
  unique <- plyr::ddply(data, c("group", "x"), uniquecols)
  unique$y <- NULL
  unique$n <- NULL

  merge(summary, unique, by = c("x", "group"), sort = FALSE)
}

uniquecols <- ggplot2:::uniquecols

bin_conf <- function(df, width = .95) {
  alpha <- (1 - width)
  results <- Hmisc::binconf(
    sum(df$y),
    sum(df$n),
    alpha = alpha,
    return.df = TRUE)
  results <- setNames(results, c("y", "ymin", "ymax"))
  results
}


StatPropCI <- ggproto(
  "StatPropCI",
  Stat,
  required_aes = c("x", "y", "n"),
  compute_panel = function(data, scales, width = .95) {
    summarise_by_xn(data, bin_conf, width = width)
  }
)
StatPropCI

# data <- d %>%
#   rename(x = Prompt, n = SWord, y = MWord)


stat_prop_ci <- function(mapping = NULL, data = NULL, geom = "pointrange",
                         position = "identity", na.rm = FALSE, show.legend = NA,
                         inherit.aes = TRUE, width = .95, ...) {
  layer(
    stat = StatPropCI, data = data, mapping = mapping, geom = geom,
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(width = width, na.rm = na.rm, ...)
  )
}

ggplot(d) +
  aes(x = Prompt, y = MWord / SWord) +
  geom_point(
    position = position_jitter(height = .025, width = .2),
    alpha = .4) +
  stat_prop_ci(
    aes(y = MWord, n = SWord),
    color = "blue",
    geom = "pointrange")


  stat_prop_ci(aes(y = MWord, n = SWord), color = "blue",  geom = "pointrange")





data <- sample_data %>%
  group_by(Prompt) %>%
  summarise(n = SWord %>% sum, y = sum(MWord)) %>%
  rename(x = Prompt)


```

First, let's go over `stat_summary()`. This function is very flexible. There are two pieces: 

1. A function to summarize the data  (fun.data, fun.y, etc.)
2. A geom to visualize the summary result.

To apply
a summary function on each group of data, we tell specify `fun.data` when the
summary gives point and range of values or `fun.y` when the summary gives a
single estimate. There are also fun.ymin or fun.ymax but I have never had
occasion to use them.

Here are some examples with `fun.y`. We give it a function to summarize the data
at each x and then a geom to plot it. In this case, the function is `mean()` and
we plot the summary with point and then with a text label.

```{r}
ggplot(d) +
  aes(x = Item, y = MWord) +
  stat_summary(
    fun.y = mean,
    geom = "point") +
  stat_summary(
    aes(label = Prompt),
    position = position_nudge(y = .025),
    geom = "label")
```

With `fun.data`, the summary functions need to return a dataframe with the columns `y`, `ymin`, `ymax`. ggplot2 provides `mean_se()`, `mean_sdl()`, `mean_cl_normal()`, and `mean_cl_boot()`.


```{r}
ggplot(d) + 
  aes(x = Item, y = MWord) + 
  stat_summary(
    aes(color = "1. fun.data = mean_se,\ngeom = \"pointrange\" (default)"),
    position = position_nudge(x = -.10)) +
  stat_summary(
    aes(color = "2. fun.data = mean_cl_boot,\ngeom = \"errorbar\""),
    position = position_nudge(x = .10),
    fun.data = mean_cl_boot,
    geom = "errorbar",
    width = .25) + 
  labs(
    color = "Summary") +
  theme(
    legend.key.height = unit(40, "pt"))
```

### Sidenote: Bootstrapping wins here

It doesn't go past the range of the data.

```{r}
1:100 %>% 
  purrr::map_dfr(function(x) d %>% group_by(Item) %>% sample_frac(replace = TRUE) %>% summarise(mean = mean(MWord / SWord))) %>% ggplot() + 
  aes(x = Item, y = mean) + 
  geom_jitter(width = .1)
```


## The problem of binomial probabilities

Number of trials affects the uncertainty. To compute the summary, we need a
second piece of data: The number of trials.

```{r}
# visualize 10, 25, 50, 100 observations
# tidyr::expand()

```


Basically, re-write `stat_summary()` but require a `n` aesthetic.


It helps to look at what others have done.

```{r}
StatSummary$compute_group
StatSummary$compute_panel
StatSummary$compute_layer

tidybayes:::StatPointinterval
tidybayes:::StatPointinterval$compute_panel

```
