---
title: Draft post (2021-06-28)
excerpt: ''
tags: ''
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE
)
old_dir <- getwd()
```


This blog runs on Jekyll. Nowadays, we would probably use blogdown to
create a blog using R, but blogdown didn't exist when I created it. Once
Jekyll is up and running, my workflow for writing and publishing my blog is 
therefore:

1.  write RMarkdown `.Rmd` files in the `_R` directory.
2.  `knit()` those RMarkdown files, storing the output in as a `.md` in
    `_posts`.


So we pairs of input files and output files, and then output files need 
to be regenerated whenever the input files. This year, I started using the 
targets package to manage this problem. Now, whenever one of my input 
files is changed, the output is updated automatically.

Here is a sketch of how this process works. First, let's create a silly site 
in an empty directory.

```{r}
site_dir <- file.path(tempdir(), pattern = "blognarok")
dir.create(site_dir)
setwd(site_dir)
```

```{r}
old_root_dir <- knitr::opts_knit$get("root.dir")
knitr::opts_knit$set(root.dir = site_dir)
```


```{r}
getwd(site_dir)
dir.create("_R")
dir.create("_posts")

writeLines(
  glue::glue(
    "
    # Hello world!
    
    Welcome to my site!
    "
  ), 
  "_R/2021-05-11-hello.Rmd"
)

writeLines(
  glue::glue(
    "
    # Best number
    
    My favorite number is `r rpois(1, 7)`.
    "
  ), 
  "_R/2021-05-12-favorite-number.Rmd"
)


```


Which leads this setup. We need to create the corresponding pieces in `_posts`.

```{r}
fs::dir_tree()
```


Here is how we can knit one of these posts. First, we can derive the
path of the output from the path of the input.

```{r}
knit_post <- function(file_in) {
  rmd_doc_to_md_post <- function(x, out_dir = "./_posts") {
    out_file <- paste0(tools::file_path_sans_ext(basename(x)), ".md")
        file.path(out_dir, out_file)
  }
  
  knit_it <- function(file_in, file_out) {
    library(knitr)
    # render_markdown()
    knit(input = file_in, output = file_out)
    file_out
  }
  
  callr::r(
   knit_it,
    list(file_in, rmd_doc_to_md_post(file_in))
  )  
}


knit_post("./_R/2021-05-11-hello.Rmd")

cat(brio::read_file("_posts/2021-05-11-hello.md"))
```




```{r first-network, fig.width = 5, fig.height = 3, dpi = 300, fig.asp = NA}


targets::tar_script({
  library(targets)
  # library(tarchetypes)
  
  knit_post <- function(file_in) {
    rmd_doc_to_md_post <- function(x, out_dir = "./_posts") {
      out_file <- paste0(tools::file_path_sans_ext(basename(x)), ".md")
          file.path(out_dir, out_file)
    }
    
    knit_it <- function(file_in, file_out) {
      library(knitr)
      # render_markdown()
      knit(input = file_in, output = file_out)
      file_out
    }
    
    callr::r(
     knit_it,
      list(file_in, rmd_doc_to_md_post(file_in))
    )  
  }

  list(
    tar_target(
      rmd_2021_05_11,
      "./_R/2021-05-11-hello.Rmd",
      format = "file"
    )
    ,
    tar_target(
      md_2021_05_11,
      knit_post(rmd_2021_05_11),
      format = "file"
    )
  )
})

targets::tar_make()
targets::tar_visnetwork()

```






I converted my blog to use the targets package to manage its build.



I use Jekyll to build my site currently. blogdown didn't exist when I created this site, so this is where we are.

I write my posts as .Rmd files in an `_R/` folder.

```{r}
list.files("_R/")
```

The bottom of every post includes a footer which a link to the Github
source for a post and the session information in footnote.

```{r}
# readLines("./_R/_footer.Rmd")
```

So every post depends on two

```{r}

```




```{r, include = FALSE}
setwd(old_dir)
knitr::opts_knit$set(root.dir = old_root_dir)
.parent_doc <- knitr::current_input()
```
```{r, change_to_child_when_done = "_footer.Rmd"}
```
