---
title: Use `plogis()` instead of writing your own `inv_logit()` (2021-11-23)
excerpt: ''
tags: ''
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE
)
```

This post is a public service announcement: You don't need to write your 
own inverse-logit function. I'm looking at you arm 😉

```{r}
arm::invlogit
```

R provides this function for us already. It just has a funny name: `plogis()`. Indeed, look at how `boot::inv.logit()` is implemented:

```{r}
boot::inv.logit
```

That's the important of coding advice for this post: **use `plogis()` if
you want to map from logit (log-odds) scale to the probability scale**.
But now, for fun, let's unpack why this functionality is stashed away
under the name "plogis".

## The four incantations: d, p, q, r

Statistical distributions in R are usually implemented as a bundle of
four functions. They have the pattern `(prefix)(family)` where the
prefix is one of 'd', 'p', 'q', 'r' and the family is the name of the
distribution. Hence, the normal distribution in R is implemented with
`dnorm()`, `pnorm()` `qnorm()`, and `rnorm()`. 

This pattern, I would argue, is a great! When I see
`brms::rstudent_t()`, I have a good sense of what to expect: a random
number generator for the Student-t distribution. I can also expect there
to be three functions for this distribution: `brms::dstudent_t()`,
`brms::pstudent_t()`, `brms::qstudent_t()`. The standard naming scheme
and bundle of four function also describes a kind of contract for
programming. If I want to write my own implementation of the beta
distribution using the mean-precision parameterization, I should provide
the four 'd', 'p', 'q', 'r' functions.


## r: random number generation function

But what do these four functions do? Let's start with r: random number
generation. If we think of statistical distributions as random number
generators with certain mathematical features, then we need a way to
generate numbers. `rnorm(n, mean, sd)` will generate `n` random numbers
from a distribution with a given `mean` and `sd`:

```{r}
library(tidyverse)
rnorm(10, mean = 0, sd = 1)
rnorm(20, mean = 10, sd = 2)
```

Let's say we have a 100 and then 1000 of these draws. If we plot a histogram of these random draws, we can kinda see the normal distribution's characteristic bell curve.

```{r}
set.seed(20211123)
draws_norm <- tibble(
  x = rnorm(1000, mean = 0, sd = 1)
)

current_theme <- theme_minimal(base_size = 14) +
  theme(panel.grid.minor = element_blank())

ggplot(draws_norm) + 
  aes(x = x) + 
  geom_histogram(
    data = . %>% head(100) %>% mutate(n = 100),
    binwidth = .5
  ) +
  geom_histogram(
    data = . %>% mutate(n = 1000),
    binwidth = .2
  ) +
  facet_wrap("n") +
  current_theme

# 
# draws_animation = draws_norm %>% 
#   mutate(
#     batch = ntile(seq_along(x), 10),
#   ) %>% 
#   nest(data = x) %>% 
#   mutate(
#     data = purrr::accumulate(data, bind_rows)
#   ) %>% 
#   unnest(data) %>% 
#   rename(x = data)
# 
# library(gganimate)
# ggplot(draws_animation) + 
#   aes(x = x) + 
#   geom_histogram(aes(y = stat(ncount)), bins = 50) + 
#   transition_states(batch)
```

One thing to know about ggplot2 is that some plotting functions will do
a bunch of statistical work for us and provide these values as
**computed variables**. For example, a histogram produces the `count` of
observations for each bin, and this is what we see on the previous plot.
The histogram also produces the variables `ncount` (normalized counts),
`width` (bin width), and `density` (we are working to define density
right now). The documentation for plotting functions
provides these in a list under the heading "Computed variables". 

We can access these computed variables by using `stat()` inside an
`aes()` specification. In the plot below, we can scale the histograms so
that they have the same maximum bin height by using the `ncount`
computed variable.

```{r}
ggplot(draws_norm) + 
  aes(x = x) + 
  geom_histogram(
    aes(y = stat(ncount)),
    data = . %>% head(100) %>% mutate(n = 100),
    binwidth = .5
  ) +
  geom_histogram(
    aes(y = stat(ncount)),
    data = . %>% mutate(n = 1000),
    binwidth = .2
  ) +
  facet_wrap("n") + 
  ggtitle("scale: normalized counts") +
  current_theme
```

Let's scale these again. If we divide each of these heights by the sum
of the heights, then we will have the probability of a point landing
inside that bin. These probabilities will sum to 1.

```{r}
ggplot(draws_norm) + 
  aes(x = x) + 
  geom_histogram(
    aes(y = stat(ncount / sum(ncount))),
    data = . %>% head(100) %>% mutate(n = 100),
    binwidth = .5
  ) +
  geom_histogram(
    aes(y = stat(ncount / sum(ncount))),
    data = . %>% mutate(n = 1000),
    binwidth = .2
  ) +
  # geom_freqpoly(
  #   aes(y = stat(cumsum(ncount / sum(ncount)))),
  #   data = . %>% mutate(n = 100),
  #   binwidth = .5
  # ) +
  # geom_freqpoly(
  #   aes(y = stat(cumsum(ncount / sum(ncount)))),
  #   data = . %>% mutate(n = 1000),
  #   binwidth = .2
  # ) + 
  facet_wrap("n") +
  labs(y = "bin probabilities") + 
  current_theme
```

If we add up the heights of successive bins, we have now have
*cumulative probabilities*. (More precisely, we have binned empirical
cumulative probabilities.) These range from 0 to 1, and we interpret
them as the proportion of observations less than or equal to x.

```{r}
ggplot(draws_norm) + 
  aes(x = x) + 
  geom_histogram(
    aes(y = stat(ncount / sum(ncount))),
    data = . %>% head(100) %>% mutate(n = 100),
    binwidth = .5,
    # color = "white"
  ) +
  geom_histogram(
    aes(y = stat(ncount / sum(ncount))),
    data = . %>% mutate(n = 1000),
    binwidth = .2,
    # color = "white"
  ) +
  stat_bin(
    aes(y = stat(cumsum(ncount / sum(ncount)))),
    data = . %>% mutate(n = 100),
    binwidth = .5,
    geom = "step"
  ) + 
  stat_bin(
    aes(y = stat(cumsum(ncount / sum(ncount)))),
    data = . %>% mutate(n = 1000),
    binwidth = .2,
    geom = "step"
  ) + 
  facet_wrap("n") +

  facet_wrap("n") +
  labs(y = "bin probabilities") + 
  current_theme
```

## p: cumulative distribution (probability) function

These cumulative probabilities approximate one of the four fundamental
functions: the *cumulative distribution function* (CDF). They have the
prefix *p* (presumably for *probability*) and provide the probability
P(X \<= q) where X is a random variable drawn from the distribution and
q is some particular value.

The probability of a random draw from Normal(0, 1) being less than or
equal to -1 is about .16. The proportion of our random draws equal to or
below -1 matches this probability.

```{r}
pnorm(q = -1, mean = 0, sd = 1)

mean(draws_norm$x[1:1000] <= -1)
```

We can overlay `pnorm()` directly onto the above plot using
`stat_function()`. Let's also include some annotations.

```{r}
#| pnorm-labeled, fig.width = 6, fig.height = 4
ggplot(draws_norm) + 
  aes(x = x) + 
  geom_histogram(
    aes(y = stat(ncount / sum(ncount))),
    data = . %>% head(100) %>% mutate(n = 100),
    binwidth = .5,
    # color = "white"
  ) +
  geom_histogram(
    aes(y = stat(ncount / sum(ncount))),
    data = . %>% mutate(n = 1000),
    binwidth = .2,
    # color = "white"
  ) +
  stat_function(fun = pnorm, color = "darkgreen", size = 1) +
  geom_text(
    data = data.frame(n = 100),
    x = .25, y = .70, 
    label = "pnorm()", 
    hjust = 1, 
    color = "darkgreen",
    family = "consolas", 
    fontface = "bold",
    size = 5
  ) +
  geom_text(
    data = data.frame(n = 100),
    x = .75, 
    y = .60, 
    label = "cumulative\nsum of bin\nprobabilities", 
    hjust = 0, 
    size = 4.5, 
    vjust = 1
  ) +
  stat_bin(
    aes(y = stat(cumsum(ncount / sum(ncount)))),
    data = . %>% mutate(n = 100),
    binwidth = .5,
    geom = "step"
  ) + 
  stat_bin(
    aes(y = stat(cumsum(ncount / sum(ncount)))),
    data = . %>% mutate(n = 1000),
    binwidth = .2,
    geom = "step"
  ) + 
  facet_wrap("n") +
  # geom_freqpoly(
  #   aes(y = stat(cumsum(ncount / sum(ncount)))),
  #   data = . %>% mutate(n = 100),
  #   binwidth = .5
  # ) +
  # geom_freqpoly(
  #   aes(y = stat(cumsum(ncount / sum(ncount)))),
  #   data = . %>% mutate(n = 1000),
  #   binwidth = .2
  # ) + 
  facet_wrap("n") +
  labs(y = "bin probabilities") + 
  theme_minimal(base_size = 14) +
  theme(panel.grid.minor = element_blank())
```




Finally, if we divide each of the bin probabilities by the bin-widths
(accessible by using `width` inside of `stat()`), we get the *density* of each bin. We can confirm this fact by accessing the density `stat()`.

```{r}
ggplot(draws_norm) + 
  aes(x = x) + 
  geom_histogram(
    aes(y = stat(ncount / sum(ncount) / width)),
    data = . %>% head(100) %>% mutate(n = 100),
    binwidth = .5
  ) +
  geom_histogram(
    aes(y = stat(ncount / sum(ncount) / width)),
    data = . %>% mutate(n = 1000),
    binwidth = .2
  ) +
  geom_freqpoly(
    aes(y = stat(density)),
    data = . %>% head(100) %>% mutate(n = 100),
    binwidth = .5,
    color = "darkorange"
  ) +
  stat_bin(
    aes(y = stat(density)),
    data = . %>% mutate(n = 1000),
    binwidth = .2,
    color = "orange",
    geom = "line"
  ) +

  facet_wrap("n") +
  ggtitle("scale: bin densities")
```

```{r}
#| dnorm-labeled, fig.width = 6, fig.height = 4
ggplot(draws_norm) + 
  aes(x = x) + 
  stat_function(fun = dnorm, color = "darkorange", size = 1) +
  geom_step(
    aes(y = stat(ncount / sum(ncount) / width)),
    data = . %>% head(100) %>% mutate(n = 100),
    binwidth = .5,
    stat = "bin", 
    direction = "mid"
  ) +
  geom_step(
    aes(y = stat(ncount / sum(ncount) / width)),
    data = . %>% mutate(n = 1000),
    binwidth = .2,
    stat = "bin",
    direction = "mid"
  ) +
  geom_text(
    data = data.frame(n = 100),
    x = 1.1, y = .30, 
    label = "dnorm()", 
    hjust = 0, 
    color = "darkorange",
    family = "consolas", 
    fontface = "bold",
    size = 5
  ) +
  geom_text(
    data = data.frame(n = 1000),
    x = -1, 
    y = .40, 
    label = expression(
      frac(
        paste("bin probability"), 
        paste("bin width")
      )
    ), 
    parse = TRUE,
    hjust = 1, 
    size = 3.7, 
    vjust = 1
  ) +
  facet_wrap("n") +
  labs(y = "bin densities") + 
  current_theme + 
  coord_cartesian(clip = "off")

```


Now let's scale these again so to divide by the sum of the heights. Now
the heights of each of these bars will sum to 1. But each of these
heights occurs in a bin. So let's divide these heights by the bin width. I am having trouble thinking of how to describe what we have at this point.


```{r}
# # density(draws_norm$x, n = 20)$y %>% sum
# # density(draws_norm$x, n = 20)$y %>% sum
# draws_norm %>% 
#   arrange(x) %>% 
#   mutate(
#     step = ppoints(x),
#     y = dnorm(x)
#   ) %>% 
#   ggplot() + 
#     aes(x = x, y = y) + 
#     geom_point()


  # geom_freqpoly(
  #   aes(y = stat(cumsum(ncount / sum(ncount)))),
  #   data = . %>% mutate(n = 1000),
  #   binwidth = .2, color = "blue"
  # ) +
  # stat_bin(
  #   aes(y = stat(cumsum(ncount / sum(ncount)) / .2), group = n),
  #   data = . %>% mutate(n = 1000),
  #   binwidth = .2,
  #   geom = "bar"
  # ) +
  # facet_wrap("n")
```


We can use the `density` `stat()` instead to get a smoothed over these histogram heights.

```{r, eval = FALSE}
ggplot(draws_norm) + 
  aes(x = x) + 
  geom_histogram(
    aes(y = stat(ncount)),
    data = . %>% head(100) %>% mutate(n = 100),
    binwidth = .5
  ) +
  geom_histogram(
    aes(y = stat(ncount / sum(ncount) / .2)),
    data = . %>% mutate(n = 1000),
    binwidth = .2
  ) +
  stat_bin(
    aes(y = stat(density)),
    data = . %>% mutate(n = 1000),
    binwidth = .2, 
    geom = "line", 
    color = "orange"
  ) +
  facet_wrap("n")

plot(density(draws_norm$x, bw = .2))

# geom_histogram(
  #   aes(y = stat(ncount / sum(ncount))),
  #   data = . %>% mutate(n = 1000),
  #   bins = 40
  # ) +
  geom_histogram(
    aes(y = stat(ncount / sum(ncount) / .2)),
    data = . %>% mutate(n = 1000),
    binwidth = .2
  ) +
  # stat_bin(
  #   aes(y = stat(cumsum(ncount / sum(ncount) / .2))),
  #   data = . %>% mutate(n = 1000),
  #   binwidth = .2,
  #   geom ="line"
  # ) +
  stat_bin(
    aes(y = stat(density)),
    data = . %>% mutate(n = 1000),
    binwidth = .2,
    geom ="line", color = "pink"
  ) +
  stat_bin(
    aes(y = stat(cumsum(density)) ),
    data = . %>% mutate(n = 1000),
    binwidth = .2,
    geom ="line", color = "blue"
  ) +
  # stat_bin(
  #   aes(y = stat(cumsum(ndensity))),
  #   data = . %>% mutate(n = 1000),
  #   binwidth = .2,
  #   geom ="line", color = "red"
  # ) +
  stat_bin(
    aes(y = stat(ncount / sum(ncount))),
    data = . %>% mutate(n = 1000),
    binwidth = .2,
    geom ="line", color = "orange"
  ) +
  stat_bin(
    aes(y = stat(cumsum(ncount / sum(ncount)))),
    data = . %>% mutate(n = 1000),
    binwidth = .2,
    geom ="line", color = "orange"
  ) +
  # geom_histogram(
  #   aes(y = stat(ncount / sum(ncount) / .2)),
  #   data = . %>% mutate(n = 1000),
  #   binwidth = .2
  # ) +
  # geom_histogram(
  #   aes(y = stat(density)),
  #   data = . %>% mutate(n = 1000),
  #   bins = 40,
  #   fill = "blue"
  # ) +
  facet_wrap("n") +
  stat_function(fun = dnorm)

StatDensity$compute_group

```


```{r}
ggplot() + 
  stat_function(
    fun = rnorm, 
    args = list(n = 100, mean = 10, sd = 2),
    geom = "bin", 
  )

```


```{r, include = FALSE}
.parent_doc <- knitr::current_input()
```
```{r, change_to_child_when_done = "_footer.Rmd"}
```
