---
title: Scraping pronunciations from Wiktionary (2026-08-13)
excerpt: ''
tags: ''
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE
)
```

Some frequent English words have so-called *strong* and *weak* forms. The strong
form is the "citation form" of the word, or how one might say the word all by
itself or in a phrase like "the word \_\_\_\_". The weak form is how speakers
tend to actually say the word in connected speech, and this spoken form is
indeed weaker: consonants may be deleted and vowels reduced to a more neutral
/ə/ vowel.

For example, *and*.

- strong: The word "and" (*rhymes with "hand"*)
- weak: Mom and dad. ([n] *practically only an /n/ sound*) 

And *for*:

- strong: I wrote a "for" loop (*sounds like "fore"*)
- weak: It was a free-for-all ([fɚ] *"fer"*) 
- weak: F'real, f'real. (*I think that's how you would read "fr fr"*) 

Searching for "English weak forms" on a search engine brings up mostly materials
related to learning English as a second-language language, because fully
articulating a word that is normally reduced sounds unnatural.

## Existing lists of strong/weak forms

Roach (2009) provides a list of 40 words with weak forms:

```{r}
wf_roach <- c(
  "the", "a", "an",
  "and", "but", 
  "that",
  "than",
  "his", "her", "your", 
  "she", "he", "we", "you",
  "him", "her", "them", "us",
  "at", "for", "from", "of", "to", "as", 
  "some", "there",
  "can", "could", "have", "has", "had", 
  "shall", "should", "must",
  "do", "does",
  "am", "are", "was", "were"
)
```

Roach (2009) writes from the perspective of British English, so some of the
/h/-initial weak forms are [/h/
deletions](https://theonion.com/lovable-cockney-charms-neighbors-1819563965/)
(*his* -\> [ɪz]), but we'll just take the list at face value.







<!-- Moreover, a lot of these are going to be vowel reductions (*your* -> [jer] [*yr*](https://www.youtube.com/watch?v=q_wroK6bY7Y)),  -->



Roach (2009) gives apparent rules for when to use strong versus weak forms. For
example,

  - A demonstrative/place *there* is strong: *There went my keys*
  - An existential or dummy-ish *there* is more likely to be weak: *There must
    be something wrong.*

One important thing to keep in mind, therefore, is that the strong/weak
variation is going to depend on surrounded words or sentence position. I'm going
to ignore all those niceties, because I just want to make a list of words with
weak forms.

Obendorfer (1998) provides another list of words with weak forms. I don't have
access to this book, but I found [a presentation by
Makino](https://www.scribd.com/doc/43388253/Pronunciations-in-connected-speech-A-Survey-of-Weak-Forms-in-a-Spoken-Corpus-of-American-English)
that starts from the Obendorfer list and proposes a few additions.

```{r}
wf_obendorfer <- c(
  "a", "an", "the",
  "be", "been","am", "are",
  "is", "was", "were",
  "had", "has", "have",
  "did", "do", "does",
  "can", "could", "may", "must", "shall", "should", "will", "would",
  "at", "by", "for", "from", "in", "of", "on", "per", "to", "up", "with", 
  "and", "but", "nor", "or", "so", "as", "if", "than", "till", 
  "what", "when", "who", "whom", "whose", "that", 
  "he", "her", "him", "his", "it", "me", "my", "she", 
  "their", "them", "they", "us", "we", "you", "your", 
  "one", "any", "no", "some", "such", "this", 
  "sure", "just", "not", "then", "there", 
  "come", "get", "go", "said", "says", "sit", "thank", 
  "ma'am", "Saint", "sir", "time", "times", "well"
)
wf_makino_additions <- c(
  "where", "gets", "our", "most", "which"
) 
```

Here's a question: Does these lists seem complete? I think about words like
*because* or *about* which have classic reduced forms ("cuz"/"cos" and "'bout")
and have to wonder. Maybe, *because* and *about* are typologically a different
category of word: words that can be weakened but don't sound unnatural in their
full form. 

## Looking for weak forms in Wiktionary

To identify any gaps, I decided to consult Wiktionary which marks pronunciations
for some weak forms. For example, the [entry for
*for*](https://en.wiktionary.org/wiki/for) has separate pronunciations for the
strong form and for the weak form:

> - (*strong form*) enPR: fôr
>   - (*Received Pronunciation*) IPA: /fɔː/
>     - Audio (*UK*, *stressed form*): [*player*]
>   - (*General American*, *Canada*) IPA: /fɔɹ/
>     - Audio (*General American*, *stressed form*): [*player*]
>   - (*Australia*, *New Zealand*) IPA: /foː/
>     - Audio (*Australia*, *stressed form*):	[*player*]
>   - (*Scotland*) IPA: /fɔɾ/
> - (*weak form*) enPR: fər
>   - (*Received Pronunciation*, *Australia*) IPA: /fə/
>     - Audio (*UK*, *unstressed form*): [*player*]
>   - (*General American*, *Canada*) IPA: /fɚ/
>     - Audio (*California*, *unstressed form*); "wait for me": [*player*]
>   - (*New Zealand*) IPA: /fɘ/
> - Rhymes: -ɔː(ɹ)
> - Homophones: faugh (*non-rhotic*), faw (*non-rhotic*), fore, four (horse–hoarse *merger*), Thor (th-*fronting*)

There is no helpful index of words with separate strong or weak forms, so I decided to check for more.


### Preprocessing the JSON corpus

First, we download Wiktionary from [kaikki.org](https://kaikki.org/dictionary/rawdata.html). The raw file data is ~23 gigabyte .jsonl file with one line per Wiktionary entry where each entry is a single json object. This file will be too large to handle with R, clearly, so let's first trim the file down to see. The generic JSON structure of an entry is documented here on [this page](https://kaikki.org/dictionary/errors/mapping/index.html). And we can see that at a minimum, what we want are the data for the following keys from each entry:


[^1]: It is possible to get a pared-down version of the English Wiktionary (1/7th the size of the original), but the authors warn that this file is deprecated, so we proceed without using it.


```
{
  "lang",
  "word",
  "pos",
  "sounds": {
    "note", "enpr", "tags", "ipa", "audio", "homophone", "rhymes"
  }
}
```



We are going to use command-line tools to preprocess the .jsonl file. These have
the advantage of operating line-by-line meaning that we never have to process
read a whole file into memory. But for the sake of demonstration, I am going to 
use R's `shell()` function to call these tools. 

To iterate on the dictionary, I first created mini-dictionaries for testing and
demo-ing jq queries, and these samples had 1, 10, or 100,000 entries in them.

Note that `> [file]` means to dump the output of the prior step into this file:

```{r}
setwd("F:/wiktionary/")
shell("head -n 1 raw-wiktextract-data.jsonl > test000001.jsonl")
shell("head -n 10 raw-wiktextract-data.jsonl > test000010.jsonl")
shell("head -n 100000 raw-wiktextract-data.jsonl > test100000.jsonl")
files <- list(
  test000001 = "test000001.jsonl",
  test000010 = "test000010.jsonl",
  test100000 = "test100000.jsonl",
  full_corpus = "raw-wiktextract-data.jsonl",
  # this is the name of a file we will create later
  target_corpus = "english-wiktionary-pronunciations.jsonl"
)
```

We use the [`jq` program](https://jqlang.org/) for working on the data.

[^2]: As a Windows user with RTools installed, I found the easiest way to get up and running with jq was to install it in the RTools terminal with pacman:

```
# upgrade packages
pacman -Syu
# install jq
pacman -S jq
```

But check the jq site for other installation guidance.


The jq tutorial provides a lot of demos. We can see the names of the top-level records in each entry:

```{r}
# -c means "compact" or print objects to just one line of text
shell('jq -c ". | keys" test000001.jsonl')
```

From now on, I'm going to use `glue::glue()` to assemble queries into the
template `jq -c "QUERY" FILE`. I am also going to use `shQuote()` to make sure
that things are properly quoted for the shell on a Windows machine:

```{r}
query <- ". | keys"
command <- glue::glue('jq -c {shQuote(query)} {files$test000001}')
command
shell(command)
```

Here is how we would get the language, word and part of speech for
each entry, but for just the English words in the file:

```{r}
query <- "select(.lang == \"English\") | {word, lang, pos}"
command <- glue::glue('jq -c {shQuote(query)} {files$test000010}')
command
shell(command)
```

What we ultimately want are the pronunciations for each entry. The structure of the pronunciation page is encoded in a funny way where there is a record for each bullet in the pronunciation section. Sometimes that bullet is just an audio sample or a link to the word's rhymes, as shown earlier with the *for* entry quoted above.


For the sake of readability, I am going to write a wrapper function that
writes/reads the `jq` query to/from a temporary file. This indirection allows me
to write the queries in a more legible format and it avoids any `shQuote()`
issues.

```{r}
make_jq_command <- function(input_file, query, pretty_print = FALSE, output_file = NULL) {
  a <- tempfile(fileext = "jq")
  writeLines(query, a)
  flag_pp <- if (pretty_print) "" else "-c"
  ending_pipe <- if (is.null(output_file)) "" else glue::glue("> {output_file}")
  command <- glue::glue('jq {flag_pp} -f {a} {input_file} {ending_pipe}')
  command
}
```

Here is how we would get the data for the word *for* that I showed earlier:


```{r}
query <- "select(.lang == \"English\" and .word == \"for\" and .pos == \"prep\") | 
  {
    word, lang, pos, sounds: [
      .sounds[]? | {note, enpr, tags, ipa, audio, homophone, rhymes}
    ]
  } | 
  select(.sounds | length > 0)"

shell(make_jq_command(files$test100000, query, pretty_print = TRUE))
list.files()


query <- 'select(.lang == \"English\") | {word, lang, pos, sounds: [.sounds[]? | {note, enpr, tags, ipa, audio, homophone, rhymes}]}'


file
command <- glue::glue('jq -c {shQuote(query)} {files$test000001} | jq {shQuote(".")}')
command
shell(command)

```

for demo


```{r}


query <- 'select(.lang == \"English\") | 
  {
    word, lang, pos, sounds: [
      .sounds[]? | {note, enpr, tags, ipa, audio, homophone, rhymes}
    ]
  }'
a <- tempfile(fileext = ".jq")
writeLines(query, a)
readLines(a)

strr
command <- glue::glue('jq -c -f {a} {file} | jq {shQuote(".")}')
command
shell(command)



command <- glue::glue('jq -c {shQuote(query)} {file} | head -n 1 | jq {shQuote(".")}')
```

If we apply this command, we can filter out a lot of the noise from the original json data.

```{r}
query <- 'select(.lang == \"English\") | {word, lang, pos, sounds: [.sounds[]? | {note, enpr, tags, ipa, audio, homophone, rhymes}]}'

command <- glue::glue('jq -c {shQuote(query)} {file} > test25-small.jsonl')
command
shell(command)

scales::label_bytes()(file.size("test000010.jsonl"))
scales::label_bytes()(file.size("test25-small.jsonl"))
file.size("test25-small.jsonl") / file.size("test000010.jsonl") 
```


```{r}
query <- 'select(.lang == \"English\") | {word, lang, pos, sounds: [.sounds[]? | {note, enpr, tags, ipa, audio, homophone, rhymes}]}'

file_big <- "raw-wiktextract-data.jsonl"
command <- glue::glue('jq -c {shQuote(query)} {file_big} > english-sounds.jsonl')
command
shell(command)
```





```{r}
query <- "select(.lang == \"English\") | 
  {
    word, lang, pos, sounds: [
      .sounds[]? | {note, enpr, tags, ipa, audio, homophone, rhymes}
    ]
  } | 
  select(.sounds | length > 0)"

shell(make_jq_command(files$test100000, query, output_file = "out100000.jsonl"))
list.files()

file.size("out100000.jsonl") |> scales::label_bytes()()
file.size("test100000.jsonl") |> scales::label_bytes()()
file.size("out100000.jsonl") / file.size("test100000.jsonl") 
```


This command takes a while to run...

```{r, eval = FALSE}
shell(make_jq_command(files$full_corpus, query, output_file = files$target_corpus))
list.files()
```




```{r}
file.size(files$target_corpus) |>  scales::label_bytes()()
file.size(files$target_corpus) / file.size(files$full_corpus) 

```

Now we have something that we can read into memory.

Because each line of the file is a separate JSON object for an entry, we
call `jsonlite::fromJSON()` on each line separately. jsonlite will helpfully
read the sounds entry into a dataframe.

```{r}
lines <- files$target_corpus |> 
  readr::read_lines()

library(dplyr)  
data <- lines |>  
  head(10000) |> 
  lapply(jsonlite::fromJSON) |> 
  lapply(function(x) {
    r <- x$sounds
    r$word <- x$word
    r$pos <- x$pos
    r$lang <- x$lang
    r <- lapply(r, as.character)
    r <- tibble::as_tibble(r)
    r$sounds_line_id <- seq_len(nrow(r))
    r
  }) |> 
  purrr::list_rbind(names_to = "entry_id") |> 
  mutate(
    tags = tags |> lapply(paste0, collapse = ", ")
  ) |> 
  select(lang, entry_id, word, pos, sounds_line_id, everything())
data


jsonlite::read_json()

```

For speed, however, we can skip R and have duckdb ingest the file for us. This 
approach was my original strategy until I decided to work with the full corpus.

```{r}
library(dbplyr)

db <- DBI::dbConnect(duckdb::duckdb())
DBI::dbExecute(db, "INSTALL json")

query <- sql("
  WITH sounds_long AS (
    SELECT
      word,
      lang,
      pos,
      unnest(sounds) AS sound,
      generate_subscripts(sounds, 1) as index
    FROM 'english-wiktionary-pronunciations.jsonl'
  )

  SELECT
    word,
    lang,
    pos,
    index,
    sound.note,
    sound.enpr,
    array_to_string(sound.tags, ', ') AS tags,
    sound.ipa,
    sound.audio,
    sound.homophone,
    sound.rhymes
  FROM sounds_long
")

tbl_words <- tbl(db, query) 
tbl_words
```


```{r}
tbl_words |> 
  filter(word == "for")
```


Now, we can query for words that are described as "weak forms":

```{r}
tbl_words |> 
  filter(stringr::str_detect(note, "weak form"))
```

That has a lot of redundant info, so we can group by 

```{r}
tbl_words |> 
  group_by(word, lang, pos) |> 
  filter(any(stringr::str_detect(note, "weak form"), na.rm = TRUE)) |> 
  ungroup() |> 
  distinct(word) |> 
  arrange(word) |> 
  collect() |> 
  print(n = Inf)
```


That really did not help up find very many new weak forms to consider. But, 
with this resource in hand, we can also look for other strange pronunciation
notes. 

```{r}
tbl_words |> 
  filter(stringr::str_detect(note, "before")) |> 
  distinct(word, note, ipa) |> 
  arrange(word)
```

```{r}
tbl_words |> 
  filter(stringr::str_detect(note, "after")) |> 
  distinct(word, note, ipa) |> 
  arrange(word) |> 
  print(n = 100)
```

Let's filter out the ones that have to do with a narrow dialect issue:

```{r}
tbl_words |> 
  filter(stringr::str_detect(note, "before")) |> 
  filter(!stringr::str_detect(note, "æ-raising")) |> 
  distinct(word, note, ipa) |> 
  arrange(word) |> 
print(n = Inf)
```

Maybe there is something in tne tags that can help us.

```{r}
tbl_words |> 
  group_by(word, pos) |> 
  filter(
    any(stringr::str_detect(tags, "unstressed")),
    any(stringr::str_detect(tags, "stressed"))
  ) |> 
  ungroup()
```

This seems like a good lead. 

```{r}
tbl_words |> 
  group_by(word, pos) |> 
  filter(
    any(stringr::str_detect(tags, "unstressed")),
    any(stringr::str_detect(tags, "stressed"))
  ) |> 
  ungroup() |> 
  distinct(word) |> 
  arrange(word) |> 
  collect() |>
  pull(word)
```


This result is definitely much noisier but it catches some of the Roach items we didn't have above.

I also want to draw attention to the items *Saint* and *sir*. Those words were 
also part of another of list of weak forms by Rudolph (1998). 







### Fun times

Just for fun time. What words start with /sf/ besides *sphinx* and *sphere* (and derived words)?

```{r}
tbl_words |> 
  filter(stringr::str_detect(ipa, "/[ˌˈ.]?sf")) |> 
  distinct(word, pos) |> 
  group_by(word) |> 
  mutate(pos = str_flatten(pos, collapse = ", ")) |> 
  ungroup() |> 
  distinct() |> 
  print(n = Inf)
```

Is /pw/ (as in *Puerto Rico*) in very many words?

```{r}
tbl_words |> 
  filter(stringr::str_detect(ipa, "/[ˌˈ.]?pw")) |> 
  distinct(word, pos) |> 
  group_by(word) |> 
  mutate(pos = str_flatten(pos, collapse = ", ")) |> 
  ungroup() |> 
  distinct() |> 
  arrange(word) |> 
  print(n = Inf)
```

I need to turn in my Statistics Card for not thinking of *poisson*.

One less exotic consonant cluster is /Sr/ (*shrink*):

```{r}
tbl_words |> 
  filter(stringr::str_detect(ipa, "/[ˌˈ.]?ʃɹ")) |> 
  distinct(word, pos) |> 
  group_by(word) |> 
  mutate(pos = str_flatten(pos, collapse = ", ")) |> 
  ungroup() |> 
  distinct() |> 
  print(n = Inf)
```

Okay what clusters start words?

```{r}
tbl_words |> 
  filter(!is.na(ipa)) |> 
  filter(!stringr::str_detect(ipa, "-")) |> 
  mutate(
    start = ipa |> 
      stringr::str_remove_all("[\\[/.ˈˌ]") |> 
      stringr::str_sub(1, 2)
  ) |> 
  distinct(word, start) |> 
  count(start) |> 
  filter(n != 1) |> 
  arrange(n) 

tbl_words |> 
  filter(!is.na(ipa)) |> 
  filter(!stringr::str_detect(ipa, "-")) |> 
  mutate(
    start = ipa |> 
      stringr::str_remove_all("[\\[/.ˈˌ]") |> 
      stringr::str_sub(1, 2)
  ) |> 
  filter(start == "kt")


filter(startsstringr::str_detect(ipa, "/[ˌˈ.]?ʃɹ")) |> 
  distinct(word, pos) |> 
  group_by(word) |> 
  mutate(pos = str_flatten(pos, collapse = ", ")) |> 
  ungroup() |> 
  distinct() |> 
  print(n = Inf)
```





```{r}
tbl_words |>
  filter(word == "Saint") |> 
  print(n = Inf)

tbl_words |>
  filter(word == "most") |> 
  print(n = Inf)
```



This is de


```{bash}
cat "F:/wiktionary/test-100.jsonl" | jq 'select(.lang == "English") | {word, lang, pos}'
```


```{r, include = FALSE}
.parent_doc <- knitr::current_input()
```
```{r, change_to_child_when_done = "_footer.Rmd"}
```
