---
title: "Am I going crazy or is everyone computing the HPDI wrong?"
author: "TJ Mahr"
date: "June 20, 2017"
output: 
  ghdown::github_html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

The _highest (posterior) density interval_ (HPDI) for a sample of values is
narrowest interval that contains x% of the data. The 90% HPDI would be the 
smallest interval that contains 90% of the data.

What's the 90% HPDI for these values?

```{r}
library(dplyr, warn.conflicts = FALSE)

xs <- 1:20
xs
```

Using existing implementations, we get...

```{r}
xs <- 1:20

hpdi1 <- HDInterval::hdi(xs, credMass = .9)
hpdi1

hpdi2 <- coda::HPDinterval(coda::as.mcmc(xs), .9)
hpdi2
```

But these intervals contain more than 90% of the data.

```{r}
within <- function(xs, range) {
  min(range) <= xs & xs <= max(range)
}

mean(within(xs, hpdi1))
mean(within(xs, hpdi2))
```

So I rolled my own version. The algorithm for finding the HPDI involves looking
at all intervals with x% of the data, and keeping the one with the smallest
interval. That's what I do here.

```{r}
compute_hpdi <- function(xs, prob = .9) {
  x_sorted <- sort(xs)
  n <- length(xs)

  num_to_keep <- ceiling(prob * n)
  num_to_drop <- n - num_to_keep

  possible_starts <- seq(1, num_to_drop + 1, by = 1)
  # Just count down from the other end
  possible_ends <- rev(seq(from = n, length = num_to_drop + 1, by = -1))

  # Find smallest interval
  span <- x_sorted[possible_ends] - x_sorted[possible_starts]
  edge <- which.min(span)
  edges <- c(possible_starts[edge], possible_ends[edge])

  # My requirement: length of span interval must be same as number to keep.
  # Other methods produce intervals that are 1 longer.
  stopifnot(length(edges[1]:edges[2]) == num_to_keep)
  
  x_sorted[edges]
}

compute_hpdi(xs)
mean(within(xs, compute_hpdi(xs)))
```

That seems more correct to me.

## Bunch of tests where my version does best

```{r}
compare_methods <- function(xs, prob = .9) {
  x <- rlang::f_label(rlang::enquo(xs))
  hpdi1 <- HDInterval::hdi(xs, credMass = prob)
  hpdi2 <- coda::HPDinterval(coda::as.mcmc(xs), prob)
  hpdi3 <- compute_hpdi(xs, prob)
  
  tibble::tribble(
    ~ x, ~ Method, ~ lower, ~ upper, ~ coverage, ~ target,
    x, "HDInterval", min(hpdi1), max(hpdi1), mean(within(xs, hpdi1)), prob,
    x, "coda", min(hpdi2), max(hpdi2), mean(within(xs, hpdi2)), prob,
    x, "TJ", min(hpdi3), max(hpdi3), mean(within(xs, hpdi3)), prob
  ) %>% 
    mutate(span = upper - lower)
}

compare_methods(rnorm(20))
compare_methods(rnorm(200))
compare_methods(rnorm(200000))

compare_methods(iris$Sepal.Length)
compare_methods(iris$Sepal.Width)

compare_methods(mtcars$mpg)
compare_methods(mtcars$mpg, .09312)

compare_methods(mtcars$wt)
compare_methods(mtcars$wt, .5)
```


