## Two bonus plots

When I was playing with this data for this post, I made two plots because I was curious. 

First, consider the total number of Lisas that were born.

```{r}
totals %>% 
  filter(name == "Lisa") %>% 
  summarise(name = unique(name), total = sum(n))
```

*By what year, were the first half of them born?*

First, I compute a rolling total for each name and then divided by the overall
total for that name to get a proportion.

```{r}
totals <- totals %>% 
  group_by(name) %>%
  mutate(
    total_n = cumsum(n),
    prop_born = total_n / sum(n)) %>%
  ungroup()
```

The time series on the proportions shows how the name. In terms of this plot, my
question is when does each name cross the .5 threshold.

```{r}
totals %>% 
  sample_n_of(9, name) %>% 
  ggplot() +
    aes(x = year, y = prop_born) +
    geom_hline(yintercept = .5, color = "grey50", linetype = "dashed") +
    geom_line() +
    facet_wrap("name") + 
    ylab("proportion of total born")


```

For each name, we filter down to find the row where the proportion minus .5 is
closest to 0.

```{r}
crossover_years <- totals %>%
  group_by(name) %>%
  filter(abs(prop_born - .5) == min(abs(prop_born - .5))) %>%
  ungroup()

crossover_years %>% 
  filter(name == "Emma")
```

Let's visualize these crossover years.

```{r crossover-plot}
ggplot(crossover_years) +
  aes(x = year, y = prop_born) +
  ggrepel::geom_text_repel(
    aes(label = name), 
    direction = "y", 
    segment.alpha = 0) + 
  theme(
    axis.ticks.y = element_blank(), 
    axis.text.y = element_blank(), 
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    axis.title.y = element_blank()) + 
  labs(
    title = "50% of Emmas were born after 1998",
    x = "Half of all people with this name were born after this year", 
    caption = "Source: Social Security Administration. Data from 1880-2015.")
```

Another more obvious question is to ask when each name peaked. To answer this question,

```{r}
peak_year <- totals %>%
  group_by(name) %>%
  filter(n == max(n)) %>%
  ungroup() %>% 
  top_n(25, n) 

peak_year %>% 
  ggplot() +
  aes(x = year, y = n) +
  geom_line(
    aes(group = name), 
    data = totals %>% semi_join(peak_year %>% top_n(25, n), by = "name"), 
    alpha = .15) +
  ggrepel::geom_text_repel(
    aes(label = name)) +
  labs(
    title = "Peak Linda",
    x = "Year",
    y = "N births of people with this name",
    caption = "Source: Social Security Administration. Data from 1880-2015.")
```



```{r}
df <- carData::Wong
library(dplyr)
library(ggplot2)
ggplot(df) +
  aes(x = days, y = piq) +
  geom_line(aes(group = id))


gapminder::gapminder

ggplot(gapminder::gapminder) +
  aes(x = year, y = lifeExp) +
  geom_line(aes(group = country))


#
# # ggplot(totals) +
# #   aes(x = year, y = prop_born) +
# #   geom_line() +
# #   facet_wrap("name")
#
# percentile_years <- totals %>%
#   group_by(name) %>%
#   filter(abs(prop_born - .5) == min(abs(prop_born - .5))) %>%
#   ungroup()
#
# ggplot(percentile_years) +
#   aes(x = year, y = prop_born) +
#   ggrepel::geom_text_repel(aes(label = name))
#
# peak_year <- totals %>%
#   group_by(name) %>%
#   filter(n == max(n)) %>%
#   ungroup()
#
# ggplot(peak_year) +
#   aes(x = year, y = n) +
#   ggrepel::geom_text_repel(aes(label = name))
```

How many things are in dots?

  ```{r}
count_dots <- function(...) length(quos(...))
count_dots()
count_dots(1, 2, 3)

my_dots <- quos(1, 2, 3)

# it's counting just the one thing
count_dots(my_dots)

# but here it's spliced
count_dots(!!! my_dots)
```

We use `!!! dots` to splice in our `dots` values into the dots needed by `group_by()`.
```
top50 <- babynames %>%
  group_by(name) %>%
  summarise(total = sum(n)) %>%
  top_n(50, total)

group_by(name) %>%
  mutate(
    total_n = cumsum(n),
    prop_born = total_n / sum(n)) %>%
  ungroup()

```
Here's a side-note... Note that the `filter()` here does not use any of the
columns in the dataframe `data`. `group_indices(grouped) %in% sampled_groups`
returns a vector the same length as the number of rows in `data` This same
principle is how `filter(data, FALSE)` returns a 0-row version of a dataframe.

