---
title: "Checking if quosures are missing or null"
author: "[TJ Mahr](https://tjmahr.github.io/)"
output: 
  ghdown::github_html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```
I'm still figuring out the new quosure (quotation + closure) system in rlang. 
This document has some notes on a problem I had to sort out.

I was working on function that took an unquoted column name as an argument. But
I also wanted to set `NULL` as a default value if the user did not provide a
column name. 

The general idea and usage would be something like...

```{r, eval = FALSE}
my_function <- function(data, column = NULL) {
  var <- enquo(column)
  # ... do things...
}

# We want to support both of these calls
my_function(iris)
my_function(iris, Species)
```

How do I check whether or not the column is `NULL` or a column name once it has
been `enquo`'d? rlang provides several functions to check the expression inside
of a quosure.

```{r}
library(rlang)

f <- function(var) {
  v <- enquo(var)
  str(list(
    v = v,
    quo_is_missing = quo_is_missing(v),
    quo_is_null = quo_is_null(v),
    quo_is_symbol = quo_is_symbol(v),
    quo_is_symbolic = quo_is_symbolic(v),
    quo_is_lang = quo_is_lang(v)))
}


f()
f(Time)
f(NULL)

g <- function(var = NULL) {
  v <- enquo(var)
  str(list(
    v = v,
    quo_is_missing = quo_is_missing(v),
    quo_is_null = quo_is_null(v),
    quo_is_symbol = quo_is_symbol(v),
    quo_is_symbolic = quo_is_symbolic(v),
    quo_is_lang = quo_is_lang(v)))
}

g()
g(Time)
g(NULL)
```

What to do when these functions are sending quos to other functions? Unquote
them if they are being re-`enquo`'d.

Without unquoting, the inner function just re-quotes `v`. Note that the
environment `<quosure: ...>` is neither global nor empty; `v`'s environment is
the outer function.

```{r}
h_send_quo <- function(var) {
    v <- enquo(var)
    f(v)
}

h_send_quo()
h_send_quo(Time)
h_send_quo(NULL)
```

With unquoting, the inner function sees the expression and environemtn quoted 
by `v`.

```{r}
h_send_unquoted <- function(var) {
    v <- enquo(var)
    f(!! v)
}
h_send_unquoted()
h_send_unquoted(Time)
h_send_unquoted(NULL)
```

