---
title: S3
excerpt: Finding a generalization
tags:
  - r
---

In the bayesplot package, we provide functions for visualizing MCMC diagnostics. `mcmc_rhat()` and `mcmc_neff()`. 

Here's a quick demo of what they do. They take a vector of diagnostic stats,
assign certain values different colors and labels, and plot them in order.

```{r}

```


Behind the scenes, these functions follow the same basic layout.

```{r}
mcmc_rhat <- function(rhat, ..., size = NULL) {
  check_ignored_arguments(...)
  data <- mcmc_rhat_data(rhat)

  # [... code for plotting rhat values ...]
}

mcmc_neff <- function(ratio, ..., size = NULL) {
  check_ignored_arguments(...)
  data <- mcmc_neff_data(ratio)
  
  # [... code for plotting neff values ...]
}


diagnostic_data_frame <- function(x, diagnostic = c("rhat", "neff")) {
  x <- auto_name(sort(x))
  stopifnot(!anyDuplicated(names(x)))

  diagnostic <- match.arg(diagnostic)
  fun <- match.fun(paste0("factor_", diagnostic))
  d <- dplyr::data_frame(
    diagnostic = diagnostic,
    parameter = factor(seq_along(x), labels = names(x)),
    value = x,
    rating = factor(fun(x), levels = c("high", "ok", "low"))
  )

  labels <- diagnostic_color_labels[[diagnostic]]
  d$description <- as.character(labels[d$rating])

  rownames(d) <- NULL
  d
}
```

They both check the function arguments, then create a dataframe for plotting, and return a plot of the diagnostic values. The dataframe-generating functions both do the same thing, basically:

```{r}
mcmc_rhat_data <- function(rhat, ...) {
  check_ignored_arguments(...)
  diagnostic_data_frame(
    x = validate_rhat(rhat),
    diagnostic = "rhat")
}

mcmc_neff_data <- function(ratio, ...) {
  check_ignored_arguments(...)
  diagnostic_data_frame(
    x = validate_neff_ratio(ratio),
    diagnostic = "neff")
}
```

First we check the arguments to the function, validate the diagnostic values,
and have a function called `diagnostic_data_frame()` create the dataframe used
for plotting. We tell `diagnostic_data_frame()` what kind of diagnostic values
we are sending.

Finally, our internal function `diagnostic_data_frame` figures out 

```{r}
diagnostic_data_frame <- function(x, diagnostic = c("rhat", "neff")) {
  # [... code for checking names ...]
  
  # Figure out function to use to convert x into a factor
  diagnostic <- match.arg(diagnostic)
  fun <- match.fun(paste0("diagnostic_factor_", diagnostic))
  
  # [... code for making the dataframe ...]
}
```

```{r}
factor_rhat <- function(x, breaks = c(1.05, 1.1)) {
  stopifnot(is.numeric(x),
            isTRUE(all(x > 0)),
            length(breaks) == 2)
  cut(x,
      breaks = c(-Inf, breaks, Inf),
      labels = c("low", "ok", "high"),
      ordered_result = FALSE)
}

# factor neff ratio
factor_neff <- function(ratio, breaks = c(0.1, 0.5)) {
  factor_rhat(ratio, breaks = breaks)
}
```


It occurred to me that there is something not quite right about this design.
There's a _code smell_--a sign that we could do better. We have two different
code branches depending on the diagnostic statistic, and we are sending along those values.

We can do better by using S3 objects and functions. In this system, we define a new class of objects. We create a _constructor_ function to create objects of each class. This function will validate values and return a vector with an appropriate class.

```{r}
new_rhat <- function(x) {
  # Convert a 1-d arrays to a vectors
  if (is.array(x) && length(dim(x)) == 1) {
    x <- as.vector(x)
  }
  
  # Validation
  stopifnot(is.numeric(x), !is.list(x), !is.array(x))
  if (any(x < 0, na.rm = TRUE)) {
    stop("All 'rhat' values must be positive.", call. = FALSE)
  }
  
  as_rhat(x)
}

as_rhat <- function(x) {
  structure(x, class = c("rhat", "numeric"), names = names(x))
}

`[.rhat` <- function (x, i, j, drop = TRUE, ...) {
  as_rhat(NextMethod())
}

new_neff_ratio <- function(x) {
  # Convert a 1-d arrays to a vectors
  if (is.array(x) && length(dim(x)) == 1) {
    x <- as.vector(x)
  }
  
  # Validation
  stopifnot(is.numeric(x), !is.list(x), !is.array(x))
  if (any(x < 0 | x > 1, na.rm = TRUE)) {
    stop("All neff ratios must be between 0 and 1.", call. = FALSE)
  }
  
  as_neff_ratio(x)
}

as_neff_ratio <- function(x) {
  structure(x, class = c("neff_ratio", "numeric"), names = names(x))
}

`[.neff_ratio` <- function (x, i, j, drop = TRUE, ...) {
  as_neff_ratio(NextMethod())
}



new_rhat()

x <- setNames(runif(10, 1, 1.4), letters[1:10])
rhats <- new_rhat(x)
sort(rhats)
rhats[-1]
rhats["a"]



rhats <- new_rhat(rhats)
rhats <- new_neff_ratio(rhats)
sort(rhats)
str(rhats)
class(rnorm(10))
sort(new_rhat(rhats))
new_rhat(sort(rhats))
cl
rhats <- new_rhat(runif(10, 1, 1.4))
str(rhats)
diagnostic_factor(rhats)


setNames(rhats, letters[1:10])

new_rhat(as.array(1:3))

new_rhat(-1:3)
new_rhat(c("1", "2"))
new_rhat(c(TRUE))
new_rhat(list(1:3))
new_rhat(array(1:3, dim = c(1,3)))

```

Now, we can handle the `diagnostic_factor` function.

```{r}


```

```{r}
factor_rhat <- function(x, breaks = c(1.05, 1.1)) {
  stopifnot(is.numeric(x),
            isTRUE(all(x > 0)),
            length(breaks) == 2)
  cut(x,
      breaks = c(-Inf, breaks, Inf),
      labels = c("low", "ok", "high"),
      ordered_result = FALSE)
}

# factor neff ratio
factor_neff <- function(ratio, breaks = c(0.1, 0.5)) {
  factor_rhat(ratio, breaks = breaks)
}

sort.rhat <- function(...) {
  new_rhat(sort(...))
}

x <- new_rhat(rhats)
auto_name <- function(x) setNames(x, labels(x))
diagnostic_data_frame <- function(x) {
  x <- auto_name(sort(x))
  stopifnot(!anyDuplicated(names(x)))

  d <- dplyr::data_frame(
    diagnostic = class(x)[1],
    parameter = factor(seq_along(x), labels = names(x)),
    value = x,
    rating = diagnostic_factor(x))
  )

  labels <- diagnostic_color_labels[[diagnostic]]
  d$description <- as.character(labels[d$rating])

  rownames(d) <- NULL
  d
}


```


