---
title: "Why is using list() critical for .dots = setNames() uses in dplyr?"
excerpt: 
tags: 
 - dplyr
 - nonstandard evaluation
---

I wrote [an answer](https://stackoverflow.com/questions/36067533/why-is-using-list-critical-for-dots-setnames-uses-in-dplyr/36168162#36168162) about why `setNames()` shows up sometimes in standard evaluation with dplyr. 

My explanation turned into a mini-tutorial on why those standard evaluation functions have a `.dots` argument. The basic idea is that the usual variadic argument `...` is a series of expressions that get evaluated inside of the dataframe. 

```{r, message = FALSE}
library("dplyr")

# standardize and round
z_round <- . %>% scale %>% as.numeric %>% round(2)

# The two expressions defining zSL, zSW are the `...`
iris %>% 
  mutate_(zSL = ~ z_round(Sepal.Length), 
          zSW = ~ z_round(Sepal.Width)) %>%
  tbl_df
```

If we programmatically assemble or manipulate those expressions before calling `mutate_`, we can't use that `...`, because we have a _list_ of expressions, not a series of individual expressions. We use the `.dots` argument instead.

```{r}
exps <- list(
  zSL = ~ z_round(Sepal.Length), 
  zSW = ~ z_round(Sepal.Width)
)

iris %>% mutate_(exps)

iris %>% mutate_(.dots = exps) %>% tbl_df
```

