---
title: "Recent adventures with lazyeval"
excerpt:
share: false
---

The [lazyeval](https://cran.r-project.org/web/packages/lazyeval/) package is a 
tool-set for performing nonstandard evaluation in R.
Nonstandard evaluation refers to any situation where something special happens
with how user input or code is evaluated.

For example, the `library` function doesn't evaluate variables. In the 
example below, I try to trick `library` into loading a fake package called 
`evil_package` by assigning to the package name `lazyeval`. In other words, we
have the expression `lazyeval` and its value is `"evil_package"`.

```{r}
print(.packages())

lazyeval <- "evil_package"
library(lazyeval)

# The lazyeval package is loaded now.
print(.packages())
```

But this gambit doesn't work because `library` did something special: It didn't
evaluate the expression `lazyeval`. In the source code of `library`, there is a
line `package <- as.character(substitute(package))` which replaces the value of
`package` with a character version of the expression that the user wrote.

That's a simple example of nonstandard evaluation, but it's pervasive. It's why 
you never have to quote column names in dplyr or ggplot2. In this post, I
present some recent examples where I decided to use the lazyeval package to do
something nonstandard. These examples are straight out of the 
[lazyeval vignette](https://cran.r-project.org/web/packages/lazyeval/vignettes/lazyeval.html)
in terms of complexity, but that's fine. We all have to start somewhere.



## Capturing expressions

**Plot titles**. While [reading the book](https://twitter.com/tjmahr/status/764147031170551810) [_Machine Learning For Hackers_](http://shop.oreilly.com/product/0636920018483.do), I wanted
to plot random numbers generated by probability distributions discussed by the
book. I used the `lazyeval::expr_text` function to capture the command used to
generate the numbers and write it as the title of the plot.

```{r plot_dist examples, fig.height = 5, fig.width = 6}
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)

plot_dist <- function(xs) {
  data <- data_frame(x = xs)
  ggplot(data) +
    aes(x = x) +
    geom_density() +
    ggtitle(lazyeval::expr_text(xs)) 
}

plot_dist(rcauchy(n = 250, location = 0, scale = 1))
plot_dist(rgamma(n = 25000, shape = 5, rate = .5))
plot_dist(rexp(n = 25000, rate = .5))
```

**Less fussy warning messages**. I recently inherited some code where there were
custom warning messages based on the input. The code threw a warning whenever a
duplicate participant ID was found in a survey. It went something like this:

```{r}
# some dummy data
study1 <- data_frame(
  id = c(1, 2, 3, 4), 
  response = c("b", "c", "a", "b"))

study2 <- data_frame(
  id = c(1, 2, 2, 3, 1), 
  response = c("a", "a", "a", "b", "c"))

if (anyDuplicated(study1$id)) {
  warning("Duplicate IDs found in Study1", call. = FALSE)
}

if (anyDuplicated(study2$id)) {
  warning("Duplicate IDs found in Study2", call. = FALSE)
}
```

To extend this code to a new study, one would just copy-and-paste and update the
`if` statement's condition and warning messages. Like so:

```{r}
study3 <- data_frame(
  id = c(1, 2, 3, 2), 
  response = c("b", "c", "a", "b"))

if (anyDuplicated(study3$id)) {
  warning("Duplicate IDs found in Study2", call. = FALSE)
}
```

Wait, that's not right! I forgot to update the warning message...

This setup is too brittle for me, so I abstracted the procedure into a 
function. First, I wrote a helper function to print out duplicates elements in a
vector. This helper will let us make the warning message a little more
informative.

```{r}
# Print out duplicated elements in a vector
print_duplicates <- function(xs) {
  duplicated <- xs[duplicated(xs)]
  duplicated %>% sort %>% unique %>% paste0(collapse = ", ")
}

print_duplicates(study2$id)
```

Next, I wrote a function to issue the warnings. I used `lazyeval::expr_label`
convert the user-inputted expression into a string wrapped in backticks. 

```{r}
# Print a warning if duplicates are found in a vector
warn_duplicates <- function(xs) {
  if (anyDuplicated(xs)) {
    # Get what the user wrote for the xs argument
    actual_xs <- lazyeval::expr_label(xs)
    msg <- paste0("Duplicate entries in ", actual_xs, ": ",
                  print_duplicates(xs))
    warning(msg, call. = FALSE)
  }
  invisible(NULL)
}

warn_duplicates(study1$id)
warn_duplicates(study2$id)
warn_duplicates(study3$id)
```

The advantage of this approach is that the warning is a generic message that can
work on any input. But in a funny way, the warning is also customized for the
input because it includes the input printed verbatim.

An aside: In plotting, I used `lazyeval::expr_text`, but here I used 
`lazyeval::expr_label`. The two differ slighty. Namely, `expr_label` surrounds
the captured expression with backticks to indicate that expression is code:

```{r}
lazyeval::expr_text(stop())
lazyeval::expr_label(stop())
```

## Asking questions about a posterior distribution

I fit regression models with RStanARM. It lets me fit Bayesian models in 
[Stan](http://mc-stan.org/) by writing conventional R modeling code. (Btw, I'm [giving a tutorial on  RStanARM](http://www.meetup.com//MadR-Madison-R-Programming-UseRs-Group/events/233333897/?showDescription=true) in a month.)

Here's a model about some famous flowers. 

```{r, results = 'hide', cache=TRUE}
library(rstanarm)

model <- stan_glm(
  Petal.Width ~ Petal.Length * Species, 
  data = iris, 
  family = gaussian(), 
  prior = normal(0, 1))
```

Here's the essential relationship being explored.

```{r petal width height by species, fig.width = 8, fig.height = 6}
ggplot(iris) + 
  aes(x = Petal.Length, y = Petal.Width, color = Species) + 
  geom_point() + stat_smooth(method = "lm")
```

The model gives me 4000 samples from the posterior distribution of the model. 

```{r}
summary(model)
```

At the 2.5% quantile, the `Petal.Length` effect looks like zero or less than
zero. What proportion of the `Petal.Length` effects (for `setosa` flowers) is
positive?

To answer questions like this one in a convenient way, I wrote a function that a
takes boolean expression about a model's parameters and evaluates it inside of
the data-frame summary of the model posterior distribution. `lazyeval::f_eval`
does the nonstandard evaluation: It evaluates an expression captured by a
formula like `~ 0 < Petal.Length` inside of a list or data-frame.

```{r}
# Get proportion of posterior samples satisfying some inequality
posterior_proportion <- function(model, inequality) {
  draws <- as.data.frame(model)
  mean(lazyeval::f_eval(inequality, data = draws))
}

posterior_proportion(model, ~ 0 < Petal.Length)
```

Here's another question: What proportion of the posterior `Petal.Length` effect
for `virginica` flowers is positive? In classical models, you would change the 
reference level for the categorical variable, refit the model, and check the 
significance. But I don't want to refit this model because that would repeat the
MCMC sampling. (It takes about 20 seconds to fit this model after all). I'll 
just ask the model about the sum of `Petal.Length` and 
`Petal.Length:Speciesversicolor` effects. That will give me the estimated 
`Petal.Length` effect but adjusted for the `versicolor` species. 

```{r}
posterior_proportion(model, ~ 0 < Petal.Length + `Petal.Length:Speciesversicolor`)

posterior_proportion(model, ~ 0 < Petal.Length + `Petal.Length:Speciesvirginica`)
```

(The backticks around `Petal.Length:Speciesversicolor` here prevent the `:`
symbol from being evaluated as an operator.)


