---
title: "Why is using list() critical for .dots = setNames() uses in dplyr?"
excerpt: Because the functions expect a list of expressions.
tags: 
 - dplyr
 - nonstandard evaluation
 - r
 - stack exchange
---

I wrote [an answer](https://stackoverflow.com/questions/36067533/why-is-using-list-critical-for-dots-setnames-uses-in-dplyr/36168162#36168162) 
about why `setNames()` shows up sometimes in standard evaluation with dplyr. 

My explanation turned into a mini-tutorial on why those standard evaluation
functions have a `.dots` argument. The basic idea is that the usual variadic
argument `...` is a series of expressions that get evaluated inside of the
dataframe.

```{r, message = FALSE}
library(dplyr)

# standardize and round
z_round <- . %>% scale() %>% as.numeric() %>% round(2)

# The two expressions defining zMPG, zHP are the `...`
mtcars %>% 
  mutate_(
    zMPG = ~ z_round(mpg), 
    zHP = ~ z_round(hp)
  ) %>%
  as_tibble()
```

If we programmatically assemble or manipulate those expressions before calling
`mutate_()`, we can't use that `...`, because we have a *list* of expressions,
not a series of individual expressions. We use the `.dots` argument instead.

```{r}
exps <- list(
  zMPG = ~ z_round(mpg), 
  zHP = ~ z_round(hp)
)

mtcars %>% 
  mutate_(exps)

mtcars %>% 
  mutate_(.dots = exps) %>% 
  as_tibble()
```

### [*Jan. 4, 2021*] Hello from the future

Here is how you would go about this in 2021 in the tidy evaluation framework.
Use an expression-list or a quosure-list and splice them in with `!!!`.

```{r}
exprs <- rlang::exprs(
  zMPG = z_round(mpg), 
  zHP = z_round(hp)
)

# These are just expressions
exprs

mtcars %>% 
  mutate(!!! exprs) %>% 
  as_tibble()

quosures <- quos(
  zMPG = z_round(mpg), 
  zHP = z_round(hp)
)

# These are expression-environment pairs
quosures

mtcars %>% 
  mutate(!!! quosures) %>% 
  as_tibble()
```

***

*Last knitted on `r format(Sys.Date())`. [Source code on
GitHub](`r file.path(git2r::remote_url(remote = "origin"), "blob/master/_R", knitr::current_input())`).*[^si] 

[^si]: 
    ```{r, results = "markup"}
    sessioninfo::session_info()
    ```
