---
title: A tour of the tibble package 
excerpt: The tidyverse version of dataframes
tags:
  - r
---

Dataframes are used in R to hold tabular data. Think of the prototypical 
spreadsheet or database table: a grid of data arranged into rows and columns. 
That's a dataframe. The [tibble R package](http://tibble.tidyverse.org/)
provides a fresh take on dataframes to fix some longstanding annoyances with
them. For example, printing a large tibble shows just the first 10 rows instead
of the flooding the console with the first 1,000 rows.

In this post, I provide a tour of the tibble package. Because the package 
provides tools for working with tabular data, it also contain some less
well-known helper functions that I would like to advertise in this post. In
particular, I find `add_column()`, `rownames_to_column()` and
`rowid_to_column()` to be useful tools in my work.

**Why the name "tibble"?** Tibbles first appeared in the dplyr package in 
January 2014, but they weren't called "tibbles" yet. dplyr used a subclass 
`tbl_df` for its dataframe objects and they behaved like modern tibbles: Better 
printing, not converting strings to factors, etc. We loved them, and we would 
convert our plain-old dataframes into these `tbl_df`s for these features. 
However, the name tee-bee-ell-dee-eff is quite a mouthful. On Twitter,
[@JennyBryan raised the question of how to pronounce
`tbl_df`](https://twitter.com/JennyBryan/status/514652585488576512), and
[@kevin_ushey suggested "tibble 
diff"](https://twitter.com/kevin_ushey/status/514659005621219329). The name was 
enthusiastically received.
{: .notice--info}


## Creating tibbles

Create a fresh tibble using `tibble()` and vectors of values for each column.
The column definitions are evaluated sequentially, so additional columns can be 
created by manipulating earlier defined ones. Below `x` is defined and then the 
values of `x` are manipulated to create the column `x_squared`.

```{r, error = TRUE}
library(tibble)
library(magrittr)

tibble(x = 1:5, x_squared = x ^ 2)
```

Note that this sequential evaluation does not work on classical dataframes.

```{r, error = TRUE}
data.frame(x = 1:5, x_squared = x ^ 2)
```

The function `data_frame()`---note the underscore instead of a dot---is an alias
for `tibble()`, which might be more transparent if your audience has never 
heard of tibbles.

```{r}
data_frame(x = 1:5, x_squared = x ^ 2)
```

In `tibble()`, the data are defined column-by-column. We can use `tribble()` to 
write out tibbles row-by-row. Formulas like `~x` are used to denote column names.

```{r}
tribble(
  ~ Film, ~ Year,
  "A New Hope", 1977,
  "The Empire Strikes Back", 1980,
  "Return of the Jedi", 1983)
```

The name "tribble" is short for "transposed tibble" (the _transposed_
part referring to change from column-wise creation in `tibble()` to row-wise
creation in `tribble`).

I like to use light-weight tribbles for two particular tasks:

* Recoding: Create a tribble of, say, labels for a plot and join it onto a dataset.
* Exclusion: Identify observations to exclude, and remove them with an anti-join.

Pretend that we have a tibble called `dataset`. The code below shows
examples of these tasks with `dataset`.

```{r, eval = FALSE}
library(dplyr)

# Recoding example
plotting_labels <- tribble(
  ~ Group, ~ GroupLabel,
  "TD", "Typically Developing",
  "CI", "Cochlear Implant",
  "ASD", "Autism Spectrum"
)

# Attach labels to dataset
dataset <- left_join(dataset, plotting_labels, by = "Group")

# Exclusion example
ids_to_exclude <- tibble::tribble(
  ~ Study, ~ ResearchID,
  "TimePoint1", "053L",
  "TimePoint1", "102L",
  "TimePoint1", "116L"
)

reduced_dataset <- anti_join(dataset, ids_to_exclude)
```


## Converting things into tibbles

`as_tibble()` will convert dataframes, matrices, and some other types into
tibbles.

```{r}
as_tibble(mtcars)
```

We can convert simple named vectors into tibbles with `enframe()`. 
For example, `quantile()` returns a named vector which we can `enframe()`.

```{r}
quantiles <- quantile(mtcars$hp, probs = c(.1, .25, .5, .75, .9))
quantiles

enframe(quantiles, "quantile", "value")
```

I have not had an opportunity to use `enframe()` since I learned about it,
but I definitely have created dataframes from names-value pairs in the past.

It's also worth noting the most common way I create tibbles: Reading in files.
The [readr](http://readr.tidyverse.org/) package will create tibbles when 
reading in data files like csvs.



## Viewing some values from each column

When we `print()` a tibble, we only see data frame enough columns to fill the
width of the console. For example, we will not see every column in this
`tibble()`.

```{r}
# Create a 200 x 26 dataframe
df <- as.data.frame(replicate(26, 1:200)) %>% 
  setNames(letters) %>% 
  as_tibble()

df
```

We can "transpose" the printing with `glimpse()` to see a few values from every 
column. Once again, just enough data is shown to fill the width of the output
console.

```{r}
glimpse(df)
```



## Growing a tibble

We can add new rows and columns with `add_row()` and `add_column()`.

Below we add rows to the bottom of the tibble (the default behavior) and to the 
top of the tibble by using the `.before` argument (add the row _before_ row 1). 
There also is an `.after` argument, but I prefer to only add rows to the
tops and bottoms of tables. The values in the `add_row()` are computed
iteratively, so we can define values `x_squared` in terms of `x`.

```{r}
df <- tibble(comment = "original", x = 1:2, x_squared = x ^ 2)
df

df <- df %>% 
  add_row(comment = "append", x = 3:4, x_squared = x ^ 2) %>% 
  add_row(comment = "prepend", x = 0, x_squared = x ^ 2, .before = 1)

df
```

The value `NA` is used when values are not provided for a certain column.
Also, because we provide the names of the columns when adding rows, we have
don't have to write out the columns in any particular order.

```{r}
df %>% 
  add_row(x = 5, comment = "NA defaults") %>% 
  add_row(x_squared = 36, x = 6, comment = "order doesn't matter")
```

We can similarly add columns with `add_column()`.

```{r}
df %>% add_column(comment2 = "inserted column", .after = "comment")
```

Typically, with dplyr loaded, you would create new columns by using `mutate()`,
although I have recently started to prefer using `add_column()` for cases like
the above example, where I add a column with a single recycled value. 

## Row names and identifiers

Look at the converted `mtcars` tibble again.

```{r}
as_tibble(mtcars)
```

The row numbers in the converted dataframe have an asterisk `*` above them. That
means that the dataframe has row-names. Row-names are clunky and quirky; they are 
just a column of data (labels) that umm :confused: we store away from the rest
of the data. 

We should move those row-names into an explicit column, and
`rownames_to_column()` does just that.

```{r}
mtcars %>% 
  as_tibble() %>% 
  rownames_to_column("model")
```

When I fit Bayesian models, I end up with a bunch of samples from a posterior 
distribution. In my data-tidying, I need to assign a ID-number to each sample. 
The function `rowid_to_column()` automates this step by creating a new column in
a dataframe with the row-numbers. In the example below, I load some MCMC samples
from the coda package and create draw IDs.

```{r}
library(coda)
data(line, package = "coda")
line1 <- as.matrix(line$line1) %>% 
  as_tibble()

line1

line1 %>% rowid_to_column("draw")
```

From here, I could reshape the data into a long format or draw some random
samples for use in a plot, all while preserving the draw number.

*** 

...And that covers the main functionality of the tibble package. I hope you
discovered a new useful feature of the tibble package. To learn more about the
technical differences between tibbles and dataframes, see [the tibble chapter in
_R for Data Science_](http://r4ds.had.co.nz/tibbles.html).
