---
title: "10 PRINT mazes with ggplot2"
excerpt: "\\/\\/\\////\\\\/\\/\\/\\////\\/\\\\//\\/\\/\\\\\\//\\\\\\\\\\\\/\\"
tags:
  - ggplot2
  - r
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.asp = 1)
```

There is a celebrated Commodore 64 program that randomly prints outs `/`
and `\` characters and fills the screen with neat-looking maze designs.
It is just one line of code, but there is 
[a whole book](https://10print.org/) written about it.

```
10 PRINT CHR$(205.5+RND(1)); : GOTO 10
```

{% include figure image_path="/assets/images/2018-05-print10.png" alt="Screenshots of 10 PRINT in action" caption="Screenshots of the 10 PRINT program in action. Images taken from the [_10 PRINT_ book](https://10print.org/)." %}

The basic idea, from my reading of the code, is that `CHR$(205)` is `\`,
`CHR$(206)` is `/`, and the program randomly selects between the two by
adding a random number to 205.5. Endlessly looping over this command
fills the screen with that pleasing maze pattern.

In R, we could replicate this functionality with by randomly sampling
the slashes:

```{r}
set.seed(20210215)

sample_n_slashes <- function(n) {
  sample(c("/", "\\"), size = n, replace = TRUE)
}

withr::with_options(
  list(width = 40), 
  cat(sample_n_slashes(800), sep = "", fill = TRUE)
)
```

where [`withr::with_options()`][with-options] lets us temporarily
change the print width and [`cat()`][cat] concatenates the slashes
and prints out the characters as text.

We can also make this much prettier by drawing the patterns using ggplot2. 

## Drawing line segments with ggplot2

Instead of writing out slashes, we will draw a grid of diagonal line
segments, some of which will be flipped at random. To draw a segment, we
need a starting *x*–*y* coordinate and an ending *x*–*y* coordinate.
[`geom_segment()`][geom-segment] will connect the two coordinates
with a line. Here's a small example where we draw four "slashes".

```{r geom-segment-demo, fig.width = 3, out.width="50%", fig.cap="A simple demo of geom_segment()"}
library(ggplot2)
library(dplyr)

data <- tibble::tribble(
  ~row, ~col, ~x_start, ~x_end, ~y_start, ~y_end,
     1,    1,        0,      1,        0,      1,
     1,    2,        1,      2,        1,      0, # flipped
     2,    1,        0,      1,        1,      2,
     2,    2,        1,      2,        1,      2
)
  
ggplot(data) + 
  aes(x = x_start, xend = x_end, y = y_start, yend = y_end) +
  geom_segment()
```

The programming task now is to make giant grid of these slashes. Let's
start with an observation: To draw two slashes, we needed three *x*
values: 0, 1, 2. The first two served as segment starts and the last two
as segment ends. The same idea applies to the *y* values. We can
generate a bunch of starts and ends by taking a sequence of steps and
removing the first and last elements.


```{r}
# We want a 20 by 20 grid
rows <- 20
cols <- 20

x_points <- seq(0, 1, length.out = cols + 1)
x_starts <- head(x_points, -1)
x_ends <- tail(x_points, -1)

y_points <- seq(0, 1, length.out = rows + 1)
y_starts <- head(y_points, -1)
y_ends <- tail(y_points, -1)
```

Each `x_starts`--`x_ends` pair is a column in the grid, and each
`y_starts`--`y_ends` is a row in the grid. To make a slash at each
row–column combination, we have to map out all the combinations of the
rows and columns. We can do this with `crossing()` which creates all
*crossed* combinations of values. (If it helps, you might think of
*crossed* like [crossed
experiments](https://en.wikipedia.org/wiki/Factorial_experiment) or the
[Cartesian cross
product](https://en.wikipedia.org/wiki/Cartesian_product) of sets.)

```{r}
grid <- tidyr::crossing(
  # columns
  tibble(x_start = x_starts, x_end = x_ends),
  # rows
  tibble(y_start = y_starts, y_end = y_ends)
) %>%
  # So values move left to right, bottom to top
  arrange(y_start, y_end)

# 400 rows because 20 rows x 20 columns
grid
```

We can confirm that the segments in the grid fill out a plot. (I
randomly color the line segments to make individual ones visible.)

```{r grid-demo, out.width="50%", fig.cap="A grid full of line segments, all pointing the same direction"}
ggplot(grid) +
  aes(
    x = x_start, 
    y = y_start, 
    xend = x_end, 
    yend = y_end, 
    color = runif(400)
  ) +
  geom_segment(size = 1) + 
  guides(color = "none")
```

Finally, we need to flip slashes at random. A segment becomes flipped if
the `y_start` and `y_end` are swapped. In the code below, we flip the
slash in each row if a randomly drawn number between 0 and 1 is less
than .5. For style, we also use `theme_void()` to strip away the
plotting theme, leaving us with just the maze design.

```{r grid1, out.width = "50%", fig.show = 'hold', fig.align = "default", fig.cap="A maze with 50% flipping probability"}
p_flip <- .5

grid <- grid %>%
  arrange(y_start, y_end) %>% 
  mutate(
    p_flip = p_flip,
    flip = runif(length(y_end)) <= p_flip,
    y_temp1 = y_start,
    y_temp2 = y_end,
    y_start = ifelse(flip, y_temp2, y_temp1),
    y_end = ifelse(flip, y_temp1, y_temp2)
  ) %>%
  select(x_start:y_end, p_flip)

ggplot(grid) +
  aes(x = x_start, y = y_start, xend = x_end, yend = y_end) +
  geom_segment(size = 1, color = "grey20") 

last_plot() + theme_void()
```

Now, we wrap all these steps together into a pair of functions.

```{r}
make_10_print_data <- function(rows = 20, cols = 20, p_flip = .5) {
  x_points <- seq(0, 1, length.out = cols + 1)
  x_starts <- head(x_points, -1)
  x_ends <- tail(x_points, -1)
  
  y_points <- seq(0, 1, length.out = rows + 1)
  y_starts <- head(y_points, -1)
  y_ends <- tail(y_points, -1)

  grid <- tidyr::crossing(
    data.frame(x_start = x_starts, x_end = x_ends),
    data.frame(y_start = y_starts, y_end = y_ends)
  )

  grid %>%
    arrange(y_start, y_end) %>% 
    mutate(
      p_flip = p_flip,
      flip = runif(length(y_end)) <= p_flip,
      y_temp1 = y_start,
      y_temp2 = y_end,
      y_start = ifelse(flip, y_temp2, y_temp1),
      y_end = ifelse(flip, y_temp1, y_temp2)
    ) %>%
    select(x_start:y_end, p_flip)
}

draw_10_print <- function(rows = 20, cols = 20, p_flip = .5) {
  grid <- make_10_print_data(
    rows = rows, 
    cols = cols, 
    p_flip = p_flip
  )

  ggplot(grid) +
    aes(x = x_start, y = y_start, xend = x_end, yend = y_end) +
    geom_segment(size = 1, color = "grey20") 
}
```


## Now the fun part: custom flipping probabilities

We can vary the probability of flipping the slashes. For example, we can use the
density of a normal distribution to make flipping more likely for central values
and less likely for more extreme values.

```{r grid3, out.width = "50%", fig.show = 'hold', fig.align = "default", fig.cap=c("Density of the normal distribution", "A maze where flipping probability is based on the density of a normal curve")}
xs <- seq(0, 1, length.out = 40)
p_flip <- dnorm(seq(-4, 4, length.out = 40))

ggplot(data.frame(x = xs, y = p_flip)) + 
  aes(x, y) + 
  geom_line() + 
  labs(
    x = "x position", 
    y = "p(flipping)",
    title = "normal density"
  )

# We repeat p_flip for each row of the grid
draw_10_print(rows = 40, cols = 40, p_flip = rep(p_flip, 40)) + 
  theme_void()
```

We can use the cumulative density of the normal distribution so that
flipping becomes more likely as _x_ increases.

```{r grid4, out.width = "50%", fig.show = 'hold', fig.align = "default", fig.cap=c("Cumulative density of the normal distribution", "A maze where flipping probability is based on the cumulative density of a normal curve")}
xs <- seq(0, 1, length.out = 40)
p_flip <- pnorm(seq(-4, 4, length.out = 40))

ggplot(data.frame(x = xs, y = p_flip)) + 
  aes(x, y) + 
  geom_line() + 
  labs(
    x = "x position", 
    y = "p(flipping)", 
    title = "cumulative normal"
  )

draw_10_print(rows = 40, cols = 40, p_flip = rep(p_flip, 40)) + 
  theme_void()
```

The Cauchy distribution is said to have "thicker" tails than the normal
distribution, so here it shows more flips on the left and right extremes.

```{r grid-cauchy, out.width = "50%", fig.show = 'hold', fig.align = "default", fig.cap=c("Density of the Cauchy distribution", "A maze where flipping probability is based on the density of a Cauchy curve")}
xs <- seq(0, 1, length.out = 40)
p_flip <- dcauchy(seq(-4, 4, length.out = 40))

ggplot(data.frame(x = xs, y = p_flip)) + 
  aes(x, y) + 
  geom_line() + 
  labs(
    x = "x position", 
    y = "p(flipping)",
    title = "Cauchy density"
  )

draw_10_print(rows = 40, cols = 40, p_flip = rep(p_flip, 40)) + 
  theme_void()
```

The exponential distribution is a spike that quickly peters out. We can make a
probability "bowl" by splicing an exponential and a reversed exponential
together.

```{r grid5, out.width = "50%", fig.show = 'hold', fig.align = "default", fig.cap=c("Densities of two exponential curves spliced to form a bowl", "A maze where flipping probability is based on an exponential curve and a reversed exponential curve")}
# Use flipped exponential densities as probabilities
p_flip <- c(
  dexp(seq(0, 4, length.out = 20)),
  dexp(seq(4, 0, length.out = 20))
)

ggplot(data.frame(x = xs, y = p_flip)) + 
  aes(x, y) + 
  geom_line() + 
  labs(
    x = "x position", 
    y = "p(flipping)", 
    title = "exponential + flipped exponential"
  )

draw_10_print(rows = 40, cols = 40, p = rep(p_flip, 40)) +
  theme_void() 
```

We might have the probabilities increase by 10% from row to row. In the code
below, I use a simple loop to boost some random probability values by 10% from
row to row. This gives us nice streaks in the grid as a column starts to flip
for every row.

```{r grid6, fig.show = 'hold', fig.cap=c("A maze where flipping probability increases by 10% from row to row")}
boost_probs <- function(p_flip, nrows, factor = 1.1) {
  output <- p_flip
  for (i in seq_len(nrows - 1)) {
    p_flip <- p_flip * factor
    output <- c(output, p_flip)
  }
  output
}

draw_10_print(
  cols = 40, 
  rows = 40, 
  p = boost_probs(runif(40), 40, 1.1)
) +
  theme_void()
```

The probabilities can be anything we like. Here I compute the frequency of
English alphabet letters as they appear in *Pride and Prejudice* and based the
flipping probability on those values.

```{r p-p-freqs, fig.height = 3, fig.width = 5, fig.asp = 3/5, fig.cap="Letter frequencies in Pride and Prejudice"}
char_counts <- janeaustenr::prideprejudice %>% 
  tolower() %>% 
  stringr::str_split("") %>% 
  unlist() %>% 
  table()

letter_counts <- char_counts[letters] %>% as.vector()
p_letter <- letter_counts / sum(letter_counts)

ggplot(data.frame(x = letters, y = p_letter)) + 
  aes(x, y, label = x) + 
  geom_text() +
  labs(
    x = NULL, 
    y = "p(letter)", 
    title = "letter frequencies in Pride and Prejudice"
  )
```

```{r grid7, fig.height=6, fig.width = 4, fig.asp = 6/4, fig.cap="Maze where flipping frequency is based on letter frequencies in Pride and Prejudice"}
draw_10_print(cols = 26, rows = 80, p = rep(p_letter, 80)) +
  theme_void() 
```

```{r, include = FALSE}
.parent_doc <- knitr::current_input()
```
```{r, child = "_R/_footer.Rmd"}
```


[with-options]: `r downlit::autolink_url("withr::with_options()")` 
    "withr documentation"

[cat]: `r downlit::autolink_url("cat()")` 
    "cat() documentation"

[geom-segment]: `r downlit::autolink_url("ggplot2::geom_segment()")` 
    "geom_segment() documentation"
