---
title: Another mixed effects model visualization
excerpt: Creating a diagram to illustrate information borrowing (partial pooling)
tags:
  - bayesian
  - r
  - ggplot2
  - mixed effects
  - brms
  - tidybayes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  fig.asp = NULL
)
```


Last week, I presented an analysis on the longitudinal development of
intelligibility in children with cerebral palsy---that is, how well do strangers
understand these children's speech from 2 to 8 years old. My analysis used a 
Bayesian nonlinear mixed effects beta regression model. If some [models are livestock and some are
pets](https://twitter.com/hadleywickham/status/742353541793120259?s=20), this
model is my dearest pet. I first started developing it a year ago, and it took
weeks of learning and problem-solving to get the first version working
correctly. I was excited to present results from it.

But here's the thing. I couldn't just formally describe my model. Not for this
audience. My talk was at the annual conference of the American
Speech-Language-Hearing Association: The main professional gathering for
speech-language pathologists, audiologists, and researchers in communication
sciences and disorders. The audience here is rightly more concerned with
clinical or research matters than the nuts and bolts of my model.

Still, I wanted to convey the main ideas behind the model without dumbing things
down. I got to work making lots of educational diagrams. Among them was the
annotated logistic curve from 
[my earlier post](/anatomy-of-a-logistic-growth-curve/) and the following
figure, used to illustrate information borrowing (or partial pooling) in mixed
effects models:

{% include figure image_path="/assets/images/2019-11-mixed-model-diagram.png" alt="Diagram showing four panels of observed data in the top row, a set of predictions for a new set of data in the middle row, and the four panels from the first row updated to included fitted estimates." caption="Diagram I used to illustrate how mixed effects models work." %}

I am pleased with this figure. Originally, I tried to convey the idea as an
interactive process: Individual-level data feed into the population model and
those feed back into the individual estimates. I had only two sets of plots with
labeled paths running back and forth between them; it wasn't pretty. The final
plot's "feed-forward" approach simplified things a great deal. My only concern,
in hindsight, is that I should have oriented things to run left-to-right instead
of top-to-bottom so I could have filled a 16:9 widescreen slide better. (But
this vertical version probably looks great on your phone or tablet right now, so
whatever!)

In this post, I walk through how to produce a plot like this one from scratch. I
can't share the original model or the clinical data here, so I will use the
`sleepstudy` data from lme4, as in 
[my partial pooling tutorial](/plotting-partial-pooling-in-mixed-effects-models/).


```{r, include = FALSE}
library(tidyverse)
library(brms)
library(tidybayes)
library(cowplot)
```

First, let's set up our data.

```{r}
library(tidyverse)
library(brms)
library(tidybayes)
library(cowplot)

# Convert to tibble for better printing. Convert factors to strings
sleepstudy <- lme4::sleepstudy %>%
  as_tibble() %>%
  mutate(Subject = as.character(Subject))

# Add two fake participants, as in the earlier partial pooling post
df_sleep <- bind_rows(
  sleepstudy,
  tibble(Reaction = c(286, 288), Days = 0:1, Subject = "374"),
  tibble(Reaction = 245, Days = 0, Subject = "373")
)

df_sleep

# Select four participants to highlight
df_demo <- df_sleep %>%
  filter(Subject %in% c("335", "333", "350", "374"))
```

We fit a mixed model with default priors and a random-number seed for
reproducibility.

```{r code-for-display, eval = FALSE, results = "hide"}
b <- brm(
  Reaction ~ Days + (Days | Subject),
  data = df_sleep, 
  seed = 20191125
)
b
```

```{r code-for-fitting, eval = FALSE, echo = FALSE}
b <- brm(
  Reaction ~ Days + (Days | Subject),
  data = df_sleep, 
  seed = 20191125
)
readr::write_rds(b, here::here("_R/data/20191125.rds"))
```


```{r code-for-loading-and-printing, echo = FALSE}
b <- readr::read_rds(here::here("_R/data/20191125.rds"))
b
```

## Top row: connect the dots

Let's create the top row. The core of the plot is straightforward: We draw
lines, draw points, facet by `Subject`, and set a reasonable *y*-axis range for
the plot (controlling the coordinates via `coord_cartesian()`).

```{r top-row-take-1, fig.height = 2, fig.width = 6, out.width = "66%", fig.cap = "First pass at the model. It shows points connected by lines split over four panels."}
col_data <- "#2F2E33"

p_first_pass <- ggplot(df_demo) +
  aes(x = Days, y = Reaction) +
  geom_line(aes(group = Subject), color = col_data) +
  geom_point(color = col_data, size = 2) +
  facet_wrap("Subject", nrow = 1) +
  coord_cartesian(ylim = c(200, 500)) +
  ggtitle("Each participant contributes data") + 
  theme_grey(base_size = 14)

p_first_pass
```

First, let's make one somewhat obscure tweak. Currently, the left edge of the
title is aligned with the plotting panel---that is, the window with the grey
background where the data are drawn. But in our final ensemble, we are going to
have three plots and the left edge of the panel will not be in the same
location across all three plots. We want our titles to be aligned with each
other from plot to plot, so we tell ggplot2 to position the plot title using the
left edge of the `"plot"`, as opposed to the `"panel"`.

```{r top-row-take-1-tweak, fig.height = 2, fig.width = 6, out.width = "66%", fig.cap = "The previous plot with its title moved to be aligned with the left edge of the image."}
p_first_pass_tweak <- p_first_pass + 
  theme(plot.title.position = "plot")
p_first_pass_tweak
```

`r emo::ji("warning")` Note that the `plot.title.position` `theme()` option is
only available in the development version of ggplot2 as of
November 2019.
{: .notice--warning}

For the diagram, we have to remove the facet labels ("strips"),
axis titles, axis text, and axis ticks (those little lines that stick out of the
plot). We also should clean up the gridlines for the *x* axis. The *x* unit is
whole number `Days`, so putting a line ("break") at 2.5 days is not meaningful.

```{r top-row-take-2, fig.height = 2, fig.width = 6, out.width = "66%", fig.cap = "The previous plot update to have all axis text and ticks removed and have the gridlines on the panel set to even numbers."}
p_second_pass <- p_first_pass_tweak + 
  scale_x_continuous(breaks = seq(0, 9, by = 2)) +
  labs(x = NULL, y = NULL) +
  theme(
    # Removing things from `theme()` is accomplished by setting them
    # `element_blank()`
    strip.background = element_blank(),
    strip.text = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.minor = element_blank()
  )
  
p_second_pass
```

Finally, let's add the "(others)" label. Here, I use the `tag` label
as a sneaky way to add an annotation. Normally, tags are meant to label
individual plots in an ensemble display of multiple plots:

```{r tag-demo, fig.height = 2, fig.width = 3, out.width = "33%", fig.cap = "A diagram showing the placement a plot's title and tag. The tag is in the upper left corner of the plot."}
ggplot() + labs(title = "A tagged plot", tag = "A. ")
```

But we will use that feature to place some text to the right side of the plot.
Sometimes, there's a correct way and there's the way that gives you the image
you can paste into your slides, and this tag trick is one of two such shortcuts
I used in my diagram.

```{r top-row, fig.height = 2, fig.width = 6, out.width = "66%",  fig.cap = "The current plot updated to have text '(others)' added to the right of the plot to indicate the more panels exist but are omitted."}
tag_others <- "   (others)   "

p_second_pass + 
  labs(tag = tag_others) + 
  theme(
    plot.tag.position = "right",
    plot.tag = element_text(size = rel(.9))
  )
```

I have discovered that incrementally and didactically building up my plots over
multiple code chunks creates a hassle for Future Me when copypasting plotting
code, so the final, complete code is provided below.

```{r, fig.height = 2, fig.width = 6, out.width = "66%"}
col_data <- "#2F2E33"
tag_others <- "   (others)   "

p_top <- ggplot(df_demo) +
  aes(x = Days, y = Reaction) +
  geom_line(aes(group = Subject), color = col_data) +
  geom_point(color = col_data, size = 2) +
  facet_wrap("Subject", nrow = 1) +
  scale_x_continuous(breaks = seq(0, 9, by = 2)) +
  coord_cartesian(ylim = c(200, 500)) +
  ggtitle("Each participant contributes data") +
  labs(x = NULL, y = NULL, tag = tag_others) + 
  theme_grey(base_size = 14) +
  theme(
    strip.background = element_blank(),
    strip.text = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.minor = element_blank(),
    plot.tag.position = "right",
    plot.tag = element_text(size = rel(.9)),
    plot.title.position = "plot"
  )
```

## Bottom row: Beams of light

Let's skip to the bottom row because it uses the same coordinates, points and
theming as the top row. The plot visualizes the posterior fits (the estimated
mean) as a median and 95% interval.
[tidybayes](https://github.com/mjskay/tidybayes) makes the process 
straightforward with `add_fitted_draws()` to add model fits onto a dataframe and with
`stat_lineribbon()` to plot a line and ribbon summary of a posterior
distribution.

```{r}
df_demo_fitted <- df_demo %>%
  # Create a dataframe with all possible combination of Subject and Days
  tidyr::expand(
    Subject,
    Days = seq(min(Days), max(Days), by = 1)
  ) %>%
  # Get the posterior predictions
  add_fitted_draws(model = b)
```

Now we have 4,000 posterior samples for the fitted `Reaction` for each
`Subject` for each day.

```{r}
df_demo_fitted
```

Given these posterior fits, we call `stat_lineribbon()` to get the median
and 95% intervals. The plotting code is otherwise the same as the last one,
except for a different title and two lines that set the fill color and hide the
fill legend.

```{r bottom-row, fig.height = 2, fig.width = 6, out.width = "66%", fig.cap = "A plot with a showing the points from the earlier plot but showing the posterior median and 95% interval band."}
col_data <- "#2F2E33"
tag_others <- "   (others)   "

p_bottom <- ggplot(df_demo_fitted) +
  aes(x = Days, y = .value) +
  # .width is the interval width
  stat_lineribbon(alpha = .4, .width = .95) +
  geom_point(
    aes(y = Reaction), 
    data = df_demo, 
    size = 2, 
    color = col_data
  ) +
  facet_wrap("Subject", nrow = 1) +
  # Use the viridis scale on the ribbon fill
  scale_color_viridis_d(aesthetics = "fill") +
  # No legend
  guides(fill = FALSE) +
  scale_x_continuous(breaks = seq(0, 9, by = 2)) +
  coord_cartesian(ylim = c(200, 500)) +
  ggtitle(
    "Individual trajectories \"borrow information\" from others"
  ) +
  labs(x = NULL, y = NULL, tag = tag_others) + 
  theme_grey(base_size = 14) +
  theme(
    strip.background = element_blank(),
    strip.text = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.minor = element_blank(),
    plot.tag.position = "right",
    plot.tag = element_text(size = rel(.9)),
    plot.title.position = "plot"
  )

p_bottom
```



## Middle row: Piles of ribbons

For the center population plot, we are going to use posterior predicted means
for a new (as yet unobserved) participant. This type of prediction incorporates
the uncertainty for the population average (i.e., the fixed effects) and the
population variation (i.e., the random effects).

We do this by defining a new participant and obtaining their posterior fitted
values. I like to use `"fake"` as the ID for the hypothetical participant.

```{r}
df_population <- df_sleep %>%
  distinct(Days) %>%
  mutate(Subject = "fake") %>%
  add_fitted_draws(b, allow_new_levels = TRUE)
```

Next, it's just a matter of using `stat_lineribbon()` again with many, many
`.width` values to recreate that center visualization.

```{r middle-attempt-1, fig.height = 4, fig.width = 5, out.width = "66%",  fig.cap = "A plot showing several differently colored probability bands to indicate which means are most probable in the population. The population median is marked with a distracting black line."}
ggplot(df_population) +
  aes(x = Days, y = .value) +
  stat_lineribbon(
    .width = c(.1, .25, .5, .6, .7, .8, .9, .95)
  ) +
  scale_x_continuous(
    "Days",
    breaks = seq(0, 9, by = 2),
    minor_breaks = NULL
  ) +
  coord_cartesian(ylim = c(200, 500)) +
  scale_y_continuous("Reaction Time") +
  scale_color_viridis_d(aesthetics = "fill") +
  guides(fill = FALSE) +
  ggtitle("Model estimates the population of participants") +
  theme_grey(base_size = 14) +
  theme(plot.title.position = "plot")
```


Actually, not quite. That median line ruins everything. It needs to go. 

I'm too lazy to figure out how to get `stat_lineribbon()` to draw just the
ribbons. Instead, note that colors in ggplot2 can be 8-digit
hex codes, where the last two digits set the transparency for the color.

```{r hex-8-demo, fig.height = 2, fig.width = 4, out.width = "50%",  fig.cap = "A diagram showing vertical lines that become increasingly transparent."}
ggplot() + 
  geom_vline(xintercept = 1, size = 4, color = "#000000FF") + 
  geom_vline(xintercept = 2, size = 4, color = "#000000CC") + 
  geom_vline(xintercept = 3, size = 4, color = "#000000AA") +  
  geom_vline(xintercept = 4, size = 4, color = "#00000077") +  
  geom_vline(xintercept = 5, size = 4, color = "#00000044") +  
  geom_vline(xintercept = 6, size = 4, color = "#00000011") + 
  ggtitle("Using 8-digit hex colors for transparency values") +
  theme(panel.grid = element_blank())
```

This is the other shortcut I used in this diagram: I tell `stat_lineribbon()` to use
a color with `00` for the final 2 digits, so that it draws the median as a
completely transparent line.

```{r better-middle, fig.height = 4, fig.width = 5, out.width = "66%",  fig.cap = "The previous population plot but with the black line at the population median removed."}
p_population <- ggplot(df_population) +
  aes(x = Days, y = .value) +
  stat_lineribbon(
    # new part
    color = "#11111100",
    .width = c(.1, .25, .5, .6, .7, .8, .9, .95)
  ) +
  scale_x_continuous(
    "Days",
    breaks = seq(0, 9, by = 2),
    minor_breaks = NULL
  ) +
  coord_cartesian(ylim = c(200, 500)) +
  scale_y_continuous("Reaction Time") +
  scale_color_viridis_d(aesthetics = "fill") +
  guides(fill = FALSE) +
  ggtitle("Model estimates the population of participants") +
  theme_grey(base_size = 14) +
  theme(plot.title.position = "plot")
p_population
```

## Mooooooo: cowplot time

Finally, we use `plot_grid()` from cowplot[^cow] to put things together.
First, I don't want the middle population plot to be as wide as the top and
bottom rows, so I first create a `plot_grid()` containing the center plot and an
empty `NULL` spacer to its right. (I recommend [this
vignette](https://wilkelab.org/cowplot/articles/plot_grid.html) for learning how
to use `plot_grid()`.)

```{r}
p_middle <- plot_grid(
  p_population, 
  NULL, 
  nrow = 1, 
  rel_widths = c(3, .5)
)
```

Now, we just stack the three on top of each other in a single column:

```{r final-diagram, fig.width = 6, fig.height = 7, fig.cap = "The final assembled diagram."}
plot_grid(
  p_top,
  p_middle,
  p_bottom,
  ncol = 1,
  rel_heights = c(1, 2, 1)
)
```


That's it! It's amazing how much work tidybayes and cowplot save us in making
these plots. In fact, without being aware of them or their capabilities, I might
have been discouraged from even trying to create this diagram.

[^cow]: As I've mentioned elsewhere on this site, I grew up on a dairy farm `r emo::ji("cow")`, so I always read *cowplot* with an agricultural reading: These are plots of space that are being fenced off and gridded together, like one might see in [an aerial shot of farmland](https://pixels.com/featured/aerial-of-agricultural-fields-james-p-blair.html). But the package author is Claus O. Wilke, so it's probably just C.O.W.'s plotting package. *Probably.* I've never asked and don't intend to. I don't want my pastoral notions dispelled.
