---
title: "Visualizing ordinal model probabilities with ribbons"
date: 2022-03-07
tags: [r, ordinal, ggplot2, visualization]
---

```{r, include = FALSE}
library(tidyverse)
library(ordinal)
```


Here is a ggplot2 recipe for plotting cumulative
probabilities from ordinal regression models.

*** 

We have `rating` scores of various wines (1--5 point scale), and during the
production, temperature (`temp`) and skin `contact` were experimentally
controlled.

```{r}
library(tidyverse)
library(ordinal)
wine <- as_tibble(wine)
wine

model <- clm(rating ~ temp * contact, data = wine)
pred_grid <- wine |>
  distinct(contact, temp)
```

`predict.clm(type = "cum.prob")` predictions provide two sets of
cumulative probabilities P(rating) ≤ rating<sub>*k*</sub> and
P(rating) \< rating<sub>*k*</sub>. The first set yields a probability
of 1 for the highest rating, and the second set have probability 0 for
the first rating.

```{r}
pred_grid |> 
  predict(model, newdata = _, type = "cum.prob") 
```

Each rating's probability is the difference between these two sets of
cumulative probabilities, so ultimately we will visualize them directly
as a ribbon between the two cumulative probabilities.

Get a long dataframe of probabilities.

```{r}
preds <- pred_grid |>
  predict(model, newdata = _, type = "cum.prob") |>
  lapply(as.data.frame) |>
  bind_rows(.id = "prob_type") |>
  cbind(pred_grid, x = _) |>
  rename_with(stringr::str_remove, pattern = "x[.]") |> 
  tidyr::pivot_longer(
    cols = c(-prob_type, -temp, -contact),
    names_to = "rating",
    values_to = "prob"
  ) |>
  tidyr::pivot_wider(
    names_from = prob_type,
    values_from = prob
  ) |>
  rename(prob_lte = cprob1, prob_lt = cprob2)
preds
```

The less-than probabilities (`prob_lt`) provide thresholds between
rating levels.

```{r thresholds}
ggplot(preds) +
  aes(x = temp) +
  geom_line(
    aes(group = rating, y = prob_lt),
    # don't draw solid line at y = 0
    data = function(x) filter(x, prob_lt != 0),
    linetype = "dashed"
  ) +
  facet_wrap("contact")
```

Next, we might fill in the areas. `geom_ribbon()` only likes continuous
*x* axes, so we note:

```{r}
preds$temp_num <- as.numeric(preds$temp)
preds |> distinct(temp_num, temp)
```

Here the two columns of cumulative probabilities pay off because one
provides the bottom of the ribbon and the other provides the top of the
ribbon.

```{r fill-area}
ggplot(preds) +
  aes(x = as.numeric(temp)) +
  geom_ribbon(
    aes(
      fill = rating, ymin = prob_lt, ymax = prob_lte
    )
  ) +
  geom_line(
    aes(group = rating, y = prob_lt),
    # don't draw solid line at y = 0
    data = function(x) filter(x, prob_lt != 0),
    linetype = "dashed"
  ) +
  facet_wrap("contact") +
  scale_x_continuous(
    breaks = 1:2, labels = c("cold", "warm"),
    expand = expansion(mult = .25)
  ) +
  scale_color_viridis_d(option = "E", aesthetics = "fill")
```


```{r fill-label-areas}
last_plot() +
  stat_summary(
    aes(
      x = 1 + as.numeric(rating) / 5 - .05,
      label = rating,
      color = rating,
      y = (prob_lte + prob_lt) / 2
    ),
    fun = median,
    geom = "text"
  ) +
  scale_color_manual(
    values = scales::viridis_pal(option = "E")(2)[c(2,2,2,1,1)]
  ) +
  guides(color = "none", fill = "none") +
  labs(x = "temperature", y = "cumulative probability")
```

### Bayesian version

I once prepared a Bayesian version of the idea. Here is that recipe:

```{r}
b_model <- brms::brm(
  rating ~ temp * contact, 
  data = wine,
  family = brms::cumulative(), 
  backend = "cmdstanr",
  chains = 1, 
  file = "_caches/2022-03-07-ordinal"
)
```

Predictions comes as `.category` probabilities, so we sum them to
compute two cumulative probabilities. But we also compute posterior
medians to get the thresholds.

```{r}
data_rvars <- pred_grid |> 
  tidybayes::add_epred_draws(b_model) |> 
  tidybayes::nest_rvars() |> 
  group_by(contact, temp, .row) |> 
  mutate(
    prob_lte = cumsum(.epred),
    # subtracting 1 from 1 to make a 0 rvar
    prob_lt = c(prob_lte[5] - 1, prob_lte[1:4])
  )

data_boundaries <- data_rvars |>
  group_by(contact, temp, .category) |>
  ggdist::median_qi(prob_lte, prob_lt)
```

Here is the trick. Hint at the uncertainty by overlaying 100 ribbons:

```{r}
data_draws_100 <- data_rvars |> 
  tidybayes::unnest_rvars() |>
  filter(.draw <= 100)
  
# create 100 geom_area() layers
layers <- data_draws_100 |>
  split(~ .draw) |>
  lapply(function(x) {
    geom_ribbon(
      aes(fill = .category, ymin = prob_lt, ymax = prob_lte),
      alpha = .01,
      data = x
    )
  })

ggplot(data_draws_100) +
  aes(x = as.numeric(temp)) +
  layers +
  geom_line(
    aes(y = prob_lte, group = .category),
    data = data_boundaries |> filter(.category != 5),
    linetype = "dashed"
  ) +
  stat_summary(
    aes(
      x = 1 + as.numeric(.category) / 5 - .05,
      label = .category,
      color = .category,
      y = (prob_lte + prob_lt) / 2
    ),
    data = data_boundaries,
    fun = median,
    geom = "text"
  ) +
  facet_wrap("contact") +
  scale_x_continuous(
    breaks = 1:2, labels = c("cold", "warm"),
    expand = expansion(mult = .25)
  ) +
  scale_color_viridis_d(option = "E", aesthetics = "fill") +
  scale_color_manual(
    values = scales::viridis_pal(option = "E")(2)[c(2,2,2,1,1)]
  ) +
  guides(color = "none", fill = "none") +
  labs(x = "temperature", y = "cumulative probability") 

```


