---
title: "Deleting or blanking a facet cell in ggplot2"
date: 2022-01-01
tags: [r, ggplot2, visualization]
---

Suppose you have some ragged data.

```{r facet-1}
set.seed(20220121)
library(tidyverse)
data <- tibble(
  id = letters[1:14],
  group = rep(c("g1", "g2", "g3"), times = c(5, 4, 5)),
  data = lapply(1:14, function(x) {
    # just making some data
    n_points <- sample(c(2:5), size = 1) 
    ages <- sort(sample(c(10:24), size = n_points))
    data.frame(age = ages, score = rbeta(n_points, 1, 4))
  })
) %>% 
  tidyr::unnest(data)

ggplot(data, aes(x = age, y = score)) + 
  geom_point() + 
  geom_line(aes(color = group)) + 
  facet_wrap("id", ncol = 5)
```

We would like to have one row per group. So we actually would want four
panels on the middle row. If we add an extra panel to fill that space and then
remove it, we can achieve this goal. 

First add the dummy panel.

```{r facet-2}
data <- data %>% 
  tibble::add_row(id = "i2", group = "g2", age = median(data$age), score = 0) %>% 
  # two points so that geom_line() doesn't warn
  tibble::add_row(id = "i2", group = "g2", age = median(data$age), score = 0)

ggplot(data, aes(x = age, y = score)) + 
  geom_point(na.rm = TRUE) + 
  geom_line(aes(color = group), na.rm = TRUE) + 
  facet_wrap("id", ncol = 5) 
```

**2026 note: This broke over time.** My *overly complicated approach*
was to manually delete the grob using the following. Finding the
`pattern` name is very tedious because you basically have to guess the
strip and panel numbers.

```{r facet-3}
p <- last_plot()
g <- ggplotGrob(p)
# guess and check
pattern <- "panel-4-3|strip-t-5-2"
to_drop <- str_which(g$layout$name, pattern)
to_keep_names <- str_subset(g$layout$name, pattern, negate = TRUE)
# delete grob data, clean up layout table
g$grobs[to_drop] <- NULL
g$layout <- g$layout[g$layout$name %in% to_keep_names, ]
grid::grid.newpage()
grid::grid.draw(g)
```

Although I am going to give better solution, I did write this entry
precisely to have this code here for my reference. 

A better approach is to **blank the cell with a ribbon**. We have to eliminate
strip titles from the theme for this approach to work.

```{r facet-4}
ggplot(data, aes(x = age, y = score)) + 
  geom_point(na.rm = TRUE) + 
  geom_line(aes(color = group), na.rm = TRUE) + 
  geom_ribbon(
    aes(ymin = -Inf, ymax = Inf),
    data = tibble(
      id = "i2", 
      age = c(-Inf, Inf), 
      group = "g2", 
      score = NA_real_
    ),
    fill = "white",
    color = "white"
  ) +
  facet_wrap("id", ncol = 5) +
  theme(strip.text.x = element_blank())
```

[Brenton on
Twitter](https://twitter.com/bmwiernik/status/1484263988137504771?s=20)
suggested numbering IDs *within* groups and using `facet_grid()`. That
works, and it also can give labeled rows.

```{r facet-5}
data <- data %>% 
  group_by(group) %>% 
  mutate(col_num = match(id, unique(id))) %>% 
  ungroup()

ggplot(data, aes(x = age, y = score)) + 
  geom_point(na.rm = TRUE) + 
  geom_line(aes(color = group), na.rm = TRUE) + 
  geom_ribbon(
    aes(ymin = -Inf, ymax = Inf),
    data = tibble(
      col_num = 5,
      age = c(-Inf, Inf), 
      group = "g2", 
      score = NA_real_
    ),
    fill = "white",
    color = "white"
  ) +
  facet_grid(group ~ col_num) +
  theme(strip.text.x = element_blank())
```
