---
title: "Making targets handle large bootstrap workflows"
date: 2024-01-01
tags: [r, targets, pipelines, bootstrapping]
---

Here is a summary of how I got targets to handle a bunch of bootstrapping:

  - transient memory
  - not letting errors on one branch stop the whole pipeline
  - `tarchetypes::tar_group_by()`
  - `format = "fst_tbl"` or `format = "qs"`
  - `tar_read(..., 1)`

My bootstraps would instantly fill up my computer's memory during a
build. So now I dump stuff out of memory with `"transient"`. (You can do
this on a by-target level too.) I also keep running the build if one
target/branch fails:

```{r, eval = FALSE}
tar_option_set(
  error = "null",
  memory = "transient"
)
```

I also split a single target into smaller ones. I declare a global
partition value like

```{r}
N_PARTITIONS <- 80
```

Then I used `tarchetypes::tar_group_by()` to tell targets to split a
giant dataframe into groups that can be dynamically branched over. Here
is some actual code in the `_targets.R` file. The first targets
defines 2000 bootstrap replicates to be split into 80
partitions/branches and the second fits the models on each
partition/branch.

```{r, eval = FALSE}
  tar_group_by(
    straps,
    data_iwpm |>
      filter(!is.na(mean_wpm)) |>
      prepare_bootstrap_data() |>
      bootstrap_by_slpg(
        times = n_boot,
        col_child_id = subject_id,
        seed = 20220621
      ) |>
      mutate(partition = assign_partition_numbers(id, N_PARTITIONS)),
    partition
  ),
  tar_target(
    models_intelligibility,
    fit_bootstrap_models(straps, "mean_intel", "intelligibility", 3, 2),
    pattern = map(straps),
    format = "qs"
  ),
# btw the targets here themselves are built serially bc fit_bootstrap_models()
# and friends use furrr to fit the models in each branch in parallel
```

But to be honest, as I'm looking at this code, I'm realizing that this
"making a giant thing and split into branches" could have been built as "make
a small table of branch IDs and grow them within each branch". One analogy
would be to read in a dozen files at once and split up the work so each
file is processed separately versus take a dozen filenames and handle
from each filename separately.

I have to use `tarchetypes::tar_group_by()` instead of
`tarchetypes::tar_group_count()` because I have multiple datasets that I
am bootstrapping and they need to be kept together within a branch.
Like, to split the following into two branches, I would want
`bootstrap_number` 1 and 2 in one branch and 3 in a separate branch so
that I can compare `a`, `b` and `c` within each bootstrap.

```
data_set,bootstrap_number
a,1
b,1
c,1
a,2
b,2
c,2
a,3
b,3
c,3
```

For serializing the data, I used `format = "qs"` on models or lists of
models and `format = "fst_tbl"` on plain, inoffensive data tables.

I also started using `targets::tar_read(blah, n)` where `n` is some
selection of branches like `1` or `c(1, 2, 4)`, to read a subset of the
overall data. This speeds up interactive development by giving me a
smaller subset of data to write code around without having to load
gigantic objects into R. 

