---
title: "Model condition number and kappa()"
date: 2025-07-01
tags: [r, regression, diagnostics, linear-algebra]
---


A colleague got the following warning when fitting a regression model in Python:

> The condition number is large, 3.11e+03. This might indicate that
> there are strong multicollinearity or other numerical problems.

We can compute this [condition
number](https://en.wikipedia.org/wiki/Condition_number) in R via
`kappa()`

```{r}
m_mpg_disp_wt_int <- lm(mpg ~ disp * wt, mtcars)
kappa(m_mpg_disp_wt_int, exact = TRUE)
```

This kappa score indicates *something* about sensitivity or stability in
the model. "The condition number bounds how much the solution *z* of a
system of equations *Az*=*c* can change, on a relative basis, when its
components *A* and *c* are changed," says [whuber on
CV](https://stats.stackexchange.com/a/168276/14825).

R computes its default kappa from the ratio of the first and last
singular values of the upper triangle of the model's QR matrix. (I read
the [source
code](https://github.com/wch/r-source/blob/637e502dbc868d39f0813a6a2a5ef0469fe8e73e/src/library/base/R/kappa.R#L97-L121).)

```{r}
zero_out_lower_tri <- function(m) { m[lower.tri(m)] <- 0; m }
d <- m_mpg_disp_wt_int$qr$qr |> 
  # there are 4 columns, so we are making a square
  _[1:4, 1:4] |>
  zero_out_lower_tri() |> 
  svd() |> 
  _$d

d[1] / d[length(d)]
```

There is a lot of mathematical indirection here but the intuition I'm
hanging onto is that if the [SVD is warping a "circle" into an
"ellipse"](https://en.wikipedia.org/wiki/Singular_value_decomposition#Intuitive_interpretations),
the ratio in kappa is the ratio of longest and shortest axes in that
ellipse.


We can reduce this condition number by (roughly) centering our predictors:

```{r}
mean(mtcars$disp)
mtcars$disp_230 <- mtcars$disp - 230 

mean(mtcars$wt)
mtcars$wt_3 <- mtcars$wt - 3 

m_mpg_disp_wt_int <- lm(mpg ~ disp_230 * wt_3, mtcars)
kappa(m_mpg_disp_wt_int, exact = TRUE)
```

For simple non-lmem models, I prefer kind of rough centering to nice round numbers so that the
intercept can be easily described, and the centering constants can be
remembered. When more exact centering is needed, I do full on
*z*-scores instead:

```{r}
mtcars$z_disp <- mtcars$disp |> scale() |> as.vector()
mtcars$z_wt <- mtcars$wt |> scale() |> as.vector()

m_mpg_disp_wt_int <- lm(mpg ~ z_disp * z_wt, mtcars)
kappa(m_mpg_disp_wt_int, exact = TRUE)
```

As we can see, this transformation made a tremendous improvement in
condition number.

*Why doesn't R raise a warning?* There are probably lots of good reasons
to not warn by default, because we are dealing with a diagnostic score
and rules of thumb for interpreting it. That said, I was not able to
find a clear answer from an R (or S) author about why R doesn't warn
about the condition number or multicollinearity.
