---
title: "Serializing rvars with qs2's qdata"
date: 2025-05-12
tags: [r, posterior, rvar, tidybayes, serialization]
---

I was surprised by how large a tibble with `posterior::rvar()` columns became
when saved with `qs2::qs_save()`. The issue seems to be attributes/pointers on
the `rvar` objects. Removing the pointers by converting to a non-rvar format or 
using `qs2::qd_save()` will result in smaller file sizes.

***

Let's make rvars of 1000 draws each for 100 rows x 2 columns of data.

```{r}
set.seed(20250512)

f <- function(n) {
  replicate(n = n, posterior::rvar(rnorm(1000))) |> 
    do.call(what = c)
}

data <- tibble::tibble(
  a = 1:100,
  b = f(100),
  c = f(100)
)
data
```

Let's save it with `save()` and `qs2::qs_save()`, which is meant to
serialize the R object's data like `save()` does but in a faster and
smarter way.

```{r}
t1 <- tempfile()
save(data, file = t1)
file.size(t1) |> scales::label_bytes()()

t2 <- tempfile()
qs2::qs_save(data, t2, nthreads = 8)
file.size(t2) |> scales::label_bytes()()
```

If we unnest the data so that there is one row per draw, we get much
better sizes.

```{r}
t3 <- tempfile()
data_long <- tidybayes::unnest_rvars(data)
data_long

qs2::qs_save(data_long, t3, nthreads = 8)
file.size(t3) |> scales::label_bytes()()
```

```{r}
file.size(t2) / file.size(t3)
```

What is the bloat here? It is [cached
data](https://github.com/r-lib/vctrs/issues/1411) about the rvars. Here
is the data for one rvar:

```{r}
str(data$b |> unclass())
```

See the environment pointer?

Let's zap the caches and try again.

```{r}
data2 <- data
attr(data2$c, "cache") <- NULL
attr(data2$b, "cache") <- NULL

t4 <- tempfile()
qs2::qs_save(data2, t4, nthreads = 8)
file.size(t4) |> scales::label_bytes()()
```

That does reduce the size to near the unnested file size. qs2 provides
the `qd_` functions for a [stricter kind of
serialization](https://github.com/qsbase/qs2#the-qdata-format) that will
replace pointers with `NULL`, so we don't have to mess with the guts of
the rvars:

```{r}
t5 <- tempfile()
qs2::qd_save(data, t5, nthreads = 8)
file.size(t5) |> scales::label_bytes()()
```

Importantly, the data seems to come back fine:

```{r}
data3 <- qs2::qd_read(t5)
data3

data3$b == data2$b
```

