---
title: "Selecting models from lists and running anova()"
date: 2025-07-01
tags: [r, programming, tidyselect, modeling]
---

I was making a notebook with a list of related models, like

```{r}
m <- list(
  mpg_wt = lm(mpg ~ wt, mtcars),
  mpg_disp = lm(mpg ~ disp, mtcars),
  mpg_disp_wt = lm(mpg ~ disp + wt, mtcars),
  mpg_disp_wt_int = lm(mpg ~ disp * wt, mtcars),
  hp_wt = lm(hp ~ wt, mtcars),
  hp_disp = lm(hp ~ disp, mtcars),
  hp_disp_wt = lm(hp ~ disp + wt, mtcars),
  hp_disp_wt_int = lm(hp ~ disp * wt, mtcars)
)
```

I want to select a subset of models and `anova()` them, so I made a quick
tidyselect function and an S3 method method of lists with `anova()`:

```{r}
library(tidyverse)

list_select <- function(l, ...) {
  pos <- tidyselect::eval_select(c(...), l)
  rlang::set_names(l[pos], names(pos))
}

anova.list <- function(object) {
  do.call(anova, unname(object))
}

m |> 
  list_select(starts_with("mpg")) |> 
  list_select(matches("disp")) |> 
  anova()

anova(m$mpg_disp, m$mpg_disp_wt, m$mpg_disp_wt_int)
```

`list_select()` is extremely basic. It's based on the examples in the
`tidyselect::eval_select()` documentation. 

For `list_select()`, the closest option available from purrr (the
tidyverse package for working on lists) is `purrr::keep_at()`:

```{r}
m |> 
  purrr::keep_at(function(x) startsWith(x, "mpg")) |> 
  str(max.level = 1)
```

