---
title: "Confidence interval overlap and p-values"
date: 2025-07-01
tags: [r, statistics, inference, visualization, simulation]
---

Two 95% confidence intervals can overlap and still have a statistically
significant difference at *p* = .05. I want to refresh my memory on what
*p*-values are implied by different degrees of overlap and simulate the
rules of thumb myself.

Cumming ([2008](https://doi.org/10.1002/sim.3471)) provides a summary of
some visual rules of thumb. (This is the same author of 2014 [The New
Statistics](https://doi.org/10.1177/0956797613504966) manifesto.) Here
are the main points I wanted to recall:

  - Two 95% intervals can overlap by a small amount at *p* = .01.
  - Two 83.5% intervals will just touch at *p* = .05.
  - Two 95% intervals can overlap up to half the length of an interval
    whisker at *p* = .05.

We simulate data sets that yield a target *p*-value.

```{r}
simulate_group_data_for_given_p_value <- function(
    p_target, 
    n_a = 40, 
    n_b = NULL, 
    sd_a = 1, 
    sd_b = NULL, 
    seed = NULL
) {
  if (!is.null(seed)) withr::local_seed(seed)
  n_b <- n_b %||% n_a
  sd_b <- sd_b %||% sd_a
  data_a <- rnorm(n_a, 0, sd_a)
  data_b <- rnorm(n_b, 0, sd_b)
  
  get_p_val <- function(diff) t.test(data_a, data_b + diff)[["p.value"]]
  
  target_diff <- uniroot(
    f = function(x) get_p_val(x) - p_target,
    interval = c(0, 5 * sd_a),
    tol = p_target / 1000,
    extendInt = "yes"
  )
  data.frame(
    group = c(rep("a", n_a), rep("b", n_b)),
    value = c(data_a, data_b + target_diff$root),
    target_p_value = paste0("p = ", p_target)
  )
}
```

Test the data simulation functions.

```{r}
data1 <- simulate_group_data_for_given_p_value(.05, seed = 20250723)
t.test(value ~ group, data1) |> 
  broom::tidy() |> 
  str()

data2 <- simulate_group_data_for_given_p_value(.01, seed = 20250723)
t.test(value ~ group, data2) |> 
  broom::tidy() |> 
  str()
```

Create a grid of CIs and plot them. `Hmisc::smean.cl.normal()` is used by 
ggplot2 to compute the CI, and this function "the sample mean and
lower and upper Gaussian confidence limits based on the
*t*-distribution".

```{r cis, fig.width = 6, fig.height = 3}
library(tidyverse)
mean_ci <- function(x, width) {
  df <- ggplot2::mean_cl_normal(x, conf.int = width)
  df[[".width"]] <- width
  df
}

data <- bind_rows(
  data1, 
  data2,
   simulate_group_data_for_given_p_value(.005, seed = 20250723)
)

cis <- data |> 
  group_by(group, target_p_value) |> 
  reframe(
    ci = list(mean_ci(value, .95), mean_ci(value, .835))
  ) |> 
  tidyr::unnest(cols = ci)

ggplot(cis) + 
  aes(x = factor(.width), color = group) +
  geom_pointrange(
    aes(ymin = ymin, y = y, ymax = ymax),
    position = position_dodge(width = .25)
  ) + 
  facet_wrap(~target_p_value) +
  guides(color = "none") + 
  scale_x_discrete(
    "Confidence interval",
    labels = function(x) scales::label_percent(.1)(as.numeric(x))
  ) +
  labs(y = NULL) +
  theme_grey(base_size = 14)
```

I was also intrigued by the following remark:

> Sall [28] described an ingenious variation of overlap: Around any
> mean, draw a circle with radius equal to the margin of error of the 95
> per cent CI. Sall showed that if two such circles overlap so that they
> intersect at right angles, *p* = 0.05 for the comparison of the two
> means.

So, when *p* = .05, the tangents at the intersections are right angles??

```{r circles, fig.width = 4, fig.height = 2.85, out.width = "66%"}
data_circles <- data1 |> 
  bind_rows( 
    simulate_group_data_for_given_p_value(.10, seed = 20250723),
    simulate_group_data_for_given_p_value(.20, seed = 20250723)  
  ) |> 
  group_by(group, target_p_value) |> 
  reframe(
    ci = mean_ci(value, .95)
  ) |> 
  tidyr::unnest(cols = ci) |> 
  filter(.width == .95) |> 
  mutate(
    radius = (ymax - ymin) / 2
  ) |> 
  rowwise() |> 
  mutate(
    angle = list(seq(0, 2 * pi, length.out = 100))
  ) |> 
  ungroup() |> 
  tidyr::unnest(cols = angle) |> 
  mutate(
    circle_x = cos(angle) * radius, 
    circle_y = y + sin(angle) * radius
  )

ggplot(data_circles) +
  geom_point(aes(x = 0, y = y)) +
  geom_path(aes(group = y, x = circle_x, y = circle_y)) + 
  facet_wrap(~target_p_value) +
  coord_fixed()
```
