---
title: "The balanced cluster bootstrap and its implementation in rsample"
date: 2026-08-28
tags: [r, rsample, bootstrapping]
---

```{r, include = FALSE}
library(dplyr)
```

We want to resample repeated measures data at the cluster level, not at
the individual observation level. In the past, I have just resampled the
cluster IDs with replacement and then joined the original data to the
resampled IDs. In fact, I once wrote a helper function
[`wisclabmisc::join_to_split()`](https://wisclab.github.io/wisclabmisc/reference/join_to_split.html)
for combining resampled IDs to their parent dataset in an
[rsample](https://rsample.tidymodels.org/index.html)-based workflow.

```{r}
library(dplyr)
d <- lme4::toenail
data_ids <- unique(d[, "patientID", drop = FALSE])

data_ids |> 
  rsample::bootstraps(times = 10, apparent = TRUE) |>
  rename(splits_id = splits) |> 
  # Attach data to resampled ids
  mutate(
    data_splits = splits_id |> purrr::map(
      wisclabmisc::join_to_split,
      d,
      by = "patientID",
      validate = TRUE
    )
  )
```

But I was intrigued by this neat "balanced bootstrap" approach described
by [Deen and de Rooij
(2020)](https://doi.org/10.3758/s13428-019-01252-y):

> The balanced bootstrap can be used to ensure that every individual
> appears exactly *B* times in the bootstrap samples, in contrast to
> randomly drawing bootstrap samples from the parent sample. Davison and
> Hinkley (1997) show that the balanced bootstrap results in an
> efficiency gain.
> 
> For unbalanced longitudinal data, where some subjects have more
> measurements than others, the balanced bootstrap ensures that the
> average size of the bootstrap samples equals the (subject) sample size
> *N*. In the balanced bootstrap, rather than simply drawing at random,
> a matrix is made with *B* copies of the numbers *1* to *N*. This
> matrix is vectorized, randomly shuffled, and turned back into a matrix
> of size *N* × *B* (Gleason, 1988). Each of the columns of this latter
> matrix gives the indices of a single bootstrap sample.


The procedure for populating bootstrap replicates (what I call "straps")
with cluster IDs is straightforward, even in base R. Enumerate, repeat,
shuffle, split, and recombine:

```{r}
d <- lme4::toenail
data_ids <- unique(d[, "patientID", drop = FALSE])
n_ids <- nrow(data_ids)
b <- 10

assignment <- seq_len(n_ids) |>
  # Repeat and shuffle cluster indices
  rep(b) |> 
  sample() |> 
  # Divide the indices into each strap
  split(rep(seq_len(b), each = n_ids)) |> 
  unname() |>
  lapply(function(x) { 
    x <- data_ids[x, , drop = FALSE] 
    x[["cluster_id"]] <- seq_len(n_ids)
    merge(x, d)
  })

assignment |> head(2) |> str()
```

I've included a `cluster_id` column in the resulting straps, because
here's an important question: If Patient 10 shows up twice in a strap,
do we have Patient 10's rows repeated twice? Or do we just so happen to
have two separate patients have the same observations? Let's save that
for another day.

### An rsample version of the balanced cluster bootstrap

I'll port the above procedure to rsample. Part of the sales pitch for 
rsample is that a dataframe of resample assignments is lightweight, 
memory-wise:

```{r}
library(rsample)
bootstraps(d, times = 10) |> lobstr::obj_size()

# There is NOT a 10x increase in object size
bootstraps(d, times = 100) |> lobstr::obj_size()
```

Each row here is a few vectors of row indices along with a shallow copy of the
original data. The actual memory cost isn't incurred until the resampled data is
"materialized":

```{r}
data_example <- bootstraps(d, times = 100)
data_example |> lobstr::obj_size()

# build and store the actual resampled copies of the data
data_example$data <- data_example$splits |> lapply(analysis)
data_example |> lobstr::obj_size()
```

An rsample "split" object is a dataframe plus two vectors of row indices
indicating which data that are kept (*analysis*) and which are withheld
(*assessment*):

```{r}
split <- make_splits(
  list(analysis = 1:10, assessment = 11:20), 
  data = d[1:20, ]
)
split

# data plus two vectors
str(split)
```

The `analysis()` and `assessment()` methods are what actually generate 
the resulting replicates:

```{r}
analysis(split)
```

My idea for implementing a balanced cluster bootstrap would be to do the
balanced bootstrapping manually as above but also stash the full parent
data inside of each split object. Here is a version of the function with
a couple of bells and whistles:

```{r}
balanced_cluster_bootstraps <- function(data, cluster_vars, times, apparent) {
  # Helper for naming the straps
  zero_pad <- function(xs, prefix = "", width = 0) {
    # use widest element if bigger than `width`
    width <- max(c(nchar(xs), width))
    sprintf(paste0(prefix, "%0", width, "d"), xs)    
  }

  cols <- tidyselect::eval_select(rlang::enquo(cluster_vars), data) 
  data_ids <- unique(data[, cols, drop = FALSE])
  n_ids <- nrow(data_ids)
  ids <- seq_len(n_ids)
  
  assignments <- ids |>
    rep(times) |> 
    sample() |> 
    split(rep(seq_len(times), each = n_ids)) |> 
    unname()
  
  labels <- zero_pad(seq_len(times), prefix = "Bootstrap")
  
  if (apparent) {
    assignments <- c(assignments, list(ids))
    labels <- c(labels, "Apparent")
  }  
  
  splits <- assignments |> 
    lapply(function(xs) {
      l <- rsample::make_splits(
        list(analysis = xs, assessment = setdiff(ids, xs)), 
        data = data_ids,
        class = "cluster_id_sample"
      )
      l$data_parent <- data
      l
    })
  
  rsample::new_rset(
    splits = splits,
    id = labels,
    subclass = c("balanced_cluster_bootstrap", "rset")
  ) 
}

balanced_cluster_bootstraps(d, patientID, 10, apparent = TRUE)
```

Time for the clever part, I think. We override the `analysis()` and `assessment()`
methods and have them perform the table join for us:

```{r}
analysis.cluster_id_sample <- function(
    x, 
    data_parent = NULL, 
    name = "cluster_id", 
    ...
) {
  data <- as.data.frame(x, data = "analysis")
  join_names <- names(data)
  data[[name]] <- seq_len(nrow(data))
  if (is.null(data_parent)) {
    data_parent <- x$data_parent
  }
  merge(data, data_parent, by = join_names)
}

assessment.cluster_id_sample <- function(
    x, 
    data_parent = NULL, 
    name = "cluster_id", 
    ...
) {
  data <- as.data.frame(x, data = "assessment")
  join_names <- names(data)
  data[[name]] <- seq_len(nrow(data))
  if (is.null(data_parent)) {
    data_parent <- x$data_parent
  }
  merge(data, data_parent, by = join_names)
}

data_a <- balanced_cluster_bootstraps(d, patientID, 10, TRUE)
data_a[["data_analysis"]] <- lapply(data_a$splits, function(x) analysis(x))
data_a[["data_assessment"]] <- lapply(data_a$splits, function(x) assessment(x))
data_a
```

Due to some weird interaction of my blogging knitr setup and S3 method
look-up, I had to write `lapply(..., function(x) analysis(x))` above instead
of `lapply(..., analysis)`, but I assure you the latter version works
interactively in R.

Finally, let's do a quick check for cluster balance:

```{r}
data_a_unnested <- data_a |> 
  dplyr::select(-data_assessment) |> 
  dplyr::filter(id != "Apparent") |> 
  tidyr::unnest(cols = data_analysis)

data_a_unnested |> 
  dplyr::distinct(id, patientID, cluster_id) |> 
  nrow()

length(unique(d$patientID)) * 10
```
