---
title: "Quick hack to parse brms priors print-out"
date: 2026-09-22
tags: [r, brms, readr]
---

```{r, include = FALSE}
library(tidyverse)
library(dplyr)
library(brms)
```

At the time of writing, the `print()` function for a brms prior specification  
does some row-filling to make things more verbose and explicit. But these 
details are not available as a dataframe for the user to manipulate or wrangle.
[Weirdness about the prior dataframe has been reported on GitHub](https://github.com/paul-buerkner/brms/issues/1761).
This note describes a quick data-cleaning workaround. 

Let's make a small model.

```{r}
library(brms)

fivenum(Orange$circumference)
fivenum(Orange$age)

f <- bf(circumference ~ s(age, k = 4) + (1 | Tree))

p_user <- c(
  set_prior("normal(0, 100)", class = "sd"),
  set_prior("exponential(.1)", class = "sigma")
)
prior_user <- validate_prior(p_user, f, data = Orange)
prior_default <- get_prior(f, data = Orange)
```

Note how the default prior print-out includes `(flat)` in the description:

```{r}
print(prior_default)
```

If we try to use this prior object as a data.frame, the prior column has 
blanks in it instead:

```{r}
as.data.frame(prior_default)
```

Here are print-outs with a `"user"` `source`d prior:

```{r}
prior_user

as.data.frame(prior_user)
```

It seems like `print()` adjusts the `prior` and `source` columns so that there
are no blank values, either by using `"(flat)"` or by inheriting a missing prior
from the generic prior used for a parameter `class`. Visually, this inheritance
looks like the `prior` is carried down from a preceding row and the `source`
column gets the value `"(vectorized)"`.

The `as.data.frame()` version seems to have the bare minimum information
about the priors, just the non-flat distributions provided to the model.

### The Easy Workaround

If we inspect the `print()` method of the priors, we see a call to
`prepare_print_prior()`:

```{r}
brms:::print.brmsprior
```

This function does return a dataframe with filled rows that is suitable for 
wrangling:


```{r}
library(tidyverse)

brms:::prepare_print_prior(prior_user) |> 
  as.data.frame() 

# For example:
brms:::prepare_print_prior(prior_user) |> 
  as.data.frame() |> 
  filter(source != "(vectorized)")
```

### The Harder Workaround

We do not want to rely on private functions from the brms package. Private 
functions belong to the package, so they can break or change. So, let's try to
get the same sort of dataframe from just the `print()` results. The trick will be to 
treat the print-out as a fixed-width formatted data file. Instead of using a 
character to separate columns in a row, a fixed-width format gives each column a 
fixed character width. Parsing the columns is then a matter of breaking the row at 
different character positions.

If we look at the print output as data, we notice that each column is a chunk 
of leading space followed by a chunk of characters.

```{r}
l <- capture.output(print(prior_user, width = 10000))
l[1]

first_row <- l[1] |> 
  stringr::str_extract_all("\\s+\\w+") |> 
  unlist()

first_row |> 
  # show one per line
  print(width = 10)

col_names <- first_row |> stringr::str_trim()

widths <- nchar(first_row)
names(widths) <- col_names
widths
```

We can tell `readr::read_fwf()` to parse the print results using these 
character counts as widths:

```{r}
df <- readr::read_fwf(
  I(l), 
  readr::fwf_widths(widths), 
  skip = 1, 
  show_col_types = FALSE
)

df
```

Two things to note:

1.  This approach treats the first row (column names) as data, so we
    skip that first line.

2.  readr inferred column types, so that the blank cells from above were
    converted to `NA`s. 

Let's try this instead:

```{r}
df <- readr::read_fwf(
  I(l), 
  readr::fwf_widths(widths), 
  skip = 1, 
  show_col_types = FALSE, 
  col_types = readr::cols(.default = readr::col_character()), 
  na = character()
)
colnames(df) <- names(widths)
df
```

Putting this together into a single function:

```{r}
get_formatted_priors <- function(prior) {
  .prior_spec <- as.character(substitute(prior))
  
  l <- capture.output(print(prior, width = 10000))
  first_row <- l[1] |> 
    stringr::str_extract_all("\\s+\\w+") |> 
    unlist()
  col_names <- stringr::str_trim(first_row)
  widths <- nchar(first_row)
  names(widths) <- col_names

  df <- readr::read_fwf(
    I(l), 
    readr::fwf_widths(widths), 
    skip = 1, 
    show_col_types = FALSE, 
    col_types = readr::cols(.default = readr::col_character()), 
    na = character()
  )
  
  colnames(df) <- names(widths)
  df$.prior_spec <- .prior_spec
  df[c(".prior_spec", names(widths))]
}
```

Now we can, e.g., bundle up different sets of priors:

```{r}
bind_rows(
  prior_default |> get_formatted_priors(),
  prior_user |> get_formatted_priors()
) |> 
  filter(source != "(vectorized)") |> 
  select(where(function(x) any(x != "")))
```

This approach, it should be noted, is brittle. If brms ever changes how it 
prints out priors---for example, using a `tibble()` or abbreviating long 
column entries---then the above workaround breaks.
