---
title: "Computing and filtering pairwise differences in R and duckdb"
date: 2026-08-28
tags: [r, dplyr, duckdb]
---

```{r, include = FALSE}
library(dplyr)
```

We have listeners transcribe children's speech, and to filter out
potentially unreliable listeners, we will select the first 2 listeners
that are within 10 percentage points (.10 proportion units) of each
other.

```{r}
library(dplyr)
withr::local_seed(20260828)

data <- tibble::tibble(
  id = 1:20,
  mu = rnorm(20) |> plogis(),
  phi = round(rgamma(20, shape = 40)),
  a = floor(mu * phi),
  b = phi - a,
  y = Map(rbeta, 4, a, b)
) |> 
  tidyr::unnest(y) |> 
  group_by(id) |> 
  mutate(listener_num = seq_along(y)) |> 
  ungroup() |> 
  select(id, listener_num, y)
data
```

Let's work through the problem for a single vector of observations.
`outer()` will populate an outer product matrix from each combination of
values from two vectors. Here is a visualization of how elements from
vectors are paired off:

```{r}
x1 <- letters[1:5]
x2 <- 1:5
outer(x1, x2, FUN = paste0)
```

For pairwise differences, here are the steps in a very base R style:

- Compute pairwise differences into matrix
- Keep just the lower triangle of the matrix
- Get the row and column indices of pairs that meet the criteria
- Find the first such pair

Row and column indices correspond to data collection order (row 1 and
column 1 are listener 1). So, the *first such pair* is the first one with the
smallest maximum row/column index. For example, if pairs 1-4 and 2-3
both satisfy the criteria, 2-3 has to be first because 4 would not have
been collected yet by the time 2-3 satisfied the criteria.

```{r}
limit <- .1
xs <- data |> filter(id == 1) |> pull(y)
xs

diffs <- outer(xs, xs, FUN = "-")
diffs

pairs <- which(abs(diffs) <= limit, arr.ind = TRUE)
# Keep the lower triangle of the matrix
pairs <- pairs[pairs[, "col"] < pairs[, "row"], 1:2, drop = FALSE]
pairs

# Bc we have a lower triangle matrix, row index > col index, 
# so keep smallest row index
row_pair <- which.min(pairs[, "row"])
pairs[row_pair, ]

pair <- pairs[row_pair, , drop = TRUE] |> unname()
pair
```

Wrapping these steps into a function, we get:

```{r}
find_first_consistent_pair <- function(xs, limit = .1) {
  diffs <- outer(xs, xs, FUN = "-")
  pairs <- which(abs(diffs) <= limit, arr.ind = TRUE)
  # Keep the lower triangle of the matrix
  pairs <- pairs[pairs[, "col"] < pairs[, "row"], 1:2, drop = FALSE]
  
  if (nrow(pairs) == 0) return(c(NA_integer_, NA_integer_))
  
  row_pair <- which.min(pairs[, "row"])
  pair <- pairs[row_pair, , drop = TRUE] |> unname()
  pair
}

data_with_pairs <- data |> 
  group_by(id) |> 
  mutate(
    in_pair = listener_num %in% find_first_consistent_pair(y)
  ) |> 
  ungroup()
data_with_pairs
```

(Aside: I know I have couple of other base R approaches sitting around
my computer. I can't remember where I stashed them though. The point of
this note, in fact, is to put this bit of code somewhere more permanent than
a random R file on my machine.)

### Database version

In production, I have these intelligibility values computed from a duckdb 
database so I would like to make this computation using R code that can be 
converted to the duckdb dialect of SQL. 

First, let's spin up a duckdb database.

```{r}
db <- withr::local_db_connection(DBI::dbConnect(duckdb::duckdb()))
dplyr::copy_to(db, data, "data")
tbl(db, "data")
```

To do the outer product type of pairing, we do a self-join on the
tables, but we can tweak the joining criteria to keep just the
lower-triangle of pairs. I create a `self_left_join()` function for 
friendly `|>` piping. 

```{r}
self_left_join <- function(x, ...) left_join(x, x, ...)
group_vars <- rlang::syms(c("id"))

tbl_pairs <- tbl(db, "data") |> 
  self_left_join(
    join_by(
      # want A:B comparison. ordering removes redundant A:A and B:A comparisons
      !!! group_vars, x$listener_num < y$listener_num,
    ),
    suffix = c("_left", "_right")
  )
tbl_pairs
```

Because `listener_num_left < listener_num_right` as a result of the table join, 
the *first such pair* is the first one with the smallest `listener_num_right`:

```{r}
limit <- .1

tbl_selected_pairs <- tbl_pairs |> 
  group_by(!!! group_vars) |>
  mutate(diffs = abs(y_left - y_right)) |>
  filter(diffs <= limit) |> 
  filter(listener_num_right == min(listener_num_right, na.rm = TRUE)) |> 
  filter(listener_num_left == min(listener_num_left, na.rm = TRUE)) |> 
  select(
    !!! group_vars, listener_num_left, listener_num_right
  ) 
tbl_selected_pairs
```

The data are in a wide format right now, so we need to pivot them into a
longer shape. I am going to use duckdb's own functions (written in all
caps) to accomplish this task:

```{r}
tbl_pairs_to_keep <- tbl_selected_pairs |> 
  mutate(
    # unpivoting by nesting and unnesting values
    listener_num = LIST_VALUE(listener_num_left, listener_num_right) |>
      UNNEST()
  ) |>
  ungroup() |>
  select(!!! group_vars, listener_num)
tbl_pairs_to_keep
```

Finally, we can filter down to the desired listener ids.

```{r}
tbl_keep <- tbl(db, "data") |> 
  inner_join(tbl_pairs_to_keep, by = join_by(id, listener_num))
tbl_keep
```

Each `tbl_` here is a SQL query until we finally `collect()` the data
into R, so we can marvel at the SQL we generated:

```{r}
dplyr::show_query(tbl_keep)
```

Finally, we can check that the two versions agree:

```{r}
a <- data_with_pairs |> 
  filter(in_pair) |> 
  select(1, 2, 3) |> 
  arrange(id, listener_num)
b <- tbl_keep |> collect() |> arrange(id, listener_num)

all(a == b)
```


