---
title: "Using downlit on knitr code chunks for Jekyll"
date: 2026-04-24
tags: [r, meta, knitr, downlit, jekyll, syntax-highlighting]
---

I wanted to use `downlit` to add package/function links to code blocks in
my Jekyll site. downlit is an amazing feature that I have never seen used in 
other languages, and it's the main thing drawing me towards quarto instead of 
vintage RMarkdown.

Currently, this site is built in 2 steps: 

```
.Rmd files ---knitr---> 
    .md files ---jekyll---> 
       .html files
```

The simple approach *should* be 3 steps, by using `downlit::downlit_md_path()`:

```
.Rmd files ---knitr---> 
    .md files ---downlit---> 
       autolinked .md files ---jekyll---> 
       .html files
```

But this will not work cleanly because `downlit` inspects the structure
of the .md file using Pandoc, and Pandoc strips away YAML front
matter, which is bad because the YAML frontmatter is Jekyll metadata. I
know about this problem because I wrote the GitHub issue for it five
years ago 🤓 <https://github.com/r-lib/downlit/issues/123>.

But you know who else sees the code blocks? knitr. I added **a knitr
hook to run the chunk text through downlit**. In effect, I run downlit on 
the markdown output that knitr has assembled for each knitr chunk.

Below is my main knitting function. It's very verbose and hard-codes my
default knitr settings because it's meant to run in a separate (clean) R
session via callr. I added the `use_downlit_chunk_hook()` function to
register the chunk hook, and this hook runs the knitr chunk through
downlit.

```{r}
  knit_it <- function(path_in, path_out, path_figs, path_cache, base_url, use_downlit = use_downlit) {
    library(knitr)

    use_downlit_chunk_hook <- function() {
      old_chunk_hook <- knitr::knit_hooks$get("chunk")
      knitr::knit_hooks$set(chunk = function(x, options) {
        md <- old_chunk_hook(x, options)
        tmp_in <- tempfile(fileext = ".md")
        tmp_out <- tempfile(fileext = ".md")
        writeLines(md, tmp_in, useBytes = TRUE)

        downlit::downlit_md_path(
          in_path = tmp_in,
          out_path = tmp_out,
          format = "gfm"
        )
        paste(readLines(tmp_out, warn = FALSE, encoding = "UTF-8"), collapse = "\n")
      })
    }

    opts_knit$set(
      base.url = base_url,
      root.dir = here::here()
    )
    opts_chunk$set(
      fig.asp = 0.618,
      fig.width = 6,
      dpi = 300,
      fig.align = "center",
      out.width = "80%",
      fig.path = path_figs,
      cache.path = path_cache,
      fig.cap = "center",
      comment = "#>",
      collapse = TRUE,
      dev = "ragg_png"
    )
    render_markdown()
    if (use_downlit) use_downlit_chunk_hook()

    knit(path_in, path_out, envir = new.env(), encoding = "UTF-8")
  }

```

Two limitations I have to admit here:

  - Pandoc writes out the markdown file, so it needs to be a flavor of
    markdown it understands. Jekyll-markdown features may not survive this 
    transformation.
  - Because knitr is doing the hooking, downlit only sees the code
    blocks. Inline links for something like `dplyr::select()` are not
    available because that code is not in a knitr code chunk.
  - I don't have "clean" .md files. One nice thing about Markdown is that
    you can read it as plaintext, and it's still legible.

Right now, I am going to dogfood this pipeline on just the notes section of 
my site because they were all written outside of Jekyll.

After getting the hook working, I had to make autolinked code blocks
look nice. downlit outputs HTML code blocks like:

```html
<pre class="chroma"><code>...</code></pre>
```

My site get its syntax highlighting from some Ruby libraries, so I need
to put together some CSS rules for syntax highlighting that matched the
current color set. I (w/ ChatGPT) added the following lines to my .scss
file. These rules map downlit’s Chroma token classes onto the site’s
existing Base16 color scheme used by Rouge:

```scss
/* ==========================================================================
   downlit / chroma code blocks
   ========================================================================== */

pre.chroma {
  position: relative;
  margin-bottom: 1em;
  padding: 1em;
  overflow-x: auto;
  background: $base00;
  color: $base05;
  font-family: $monospace;
  font-size: $type-size-7;
  line-height: 1.5;
  border-radius: $border-radius;

  [dir=rtl] & {
    direction: ltr;
    text-align: start;
  }

  code {
    padding: 0;
    background: transparent;
    color: inherit;
    font-family: inherit;
    font-size: inherit;
  }
}

/* downlit::classes_chroma() */
.chroma {
  .c {
    /* COMMENT */
    color: $base04;
  }

  .kc {
    /* constant */
    color: $base0e;
  }

  .m {
    /* NUM_CONST */
    color: $base09;
  }

  .s {
    /* STR_CONST */
    color: $base0b;
  }

  .kr {
    /* special */
    color: $base0e;
  }

  .o {
    /* parens, infix */
    color: $base05;
  }

  .nv {
    /* SLOT, SYMBOL, SYMBOL_FORMALS */
    color: $base05;
  }

  .nf {
    /* NS_GET, NS_GET_INT, SYMBOL_FUNCTION_CALL, SYMBOL_PACKAGE */
    color: $base0c;
  }
}

pre.chroma a {
  color: inherit; /* use colour from syntax highlighting */
  text-decoration: underline;
  text-decoration-color: #ccc;
}
```

Here is a comparison of the syntax highlighting.

```{r}
# downlit
data <- mtcars$mpg[1]
x <- rnorm(100)
name <- "test"
```

```r
# rouge
data <- mtcars$mpg[1]
x <- rnorm(100)
name <- "test"
```


